---
title: "SSc Proteomics: Joint Synovitis Analysis"
author: "Analysis Script"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, dpi = 300)
```

## Overview

Binary comparison of serum Olink proteomics between SSc patients **with vs without joint synovitis**, adjusting for immunosuppression status.

**Analysis:**
- Binary limma: Joint_synovitis (Yes vs No) + Immunosuppression covariate
- Filter for AveExpr > 0, recalculate FDR
- Volcano plot and boxplots for significant hits

---

# 1. Libraries

```{r libraries}
# =============================================================================
# REQUIRED PACKAGES
# =============================================================================

library(arrow)          # Read parquet files
library(readxl)         # Read Excel metadata
library(here)           # Project-relative paths
library(dplyr)          # Data manipulation
library(tidyr)          # Reshaping data
library(tibble)         # rownames_to_column
library(limma)          # Differential expression analysis
library(ggplot2)        # Plotting
library(ggpubr)         # stat_compare_means
library(cowplot)        # Combine plots
library(ggrepel)        # Label repelling for volcano
```

---

# 2. Setup Output Directory

```{r setup-output}
# =============================================================================
# PURPOSE: Create output directory for results
# =============================================================================

# Get script name (without .Rmd extension) for output folder
script_name <- "PreSSc_Synovitis"  # Or set manually

# Create output directory: output/<script_name>/
output_dir <- here::here("output", script_name)
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

cat("Output directory:", output_dir, "\n")
```

---

# 3. Color Palettes

```{r colors}
# =============================================================================
# PURPOSE: Define consistent color schemes
# =============================================================================

colours_class <- c(
  "No" = "#3498DB",    # Blue - No synovitis

  "Yes" = "#E74C3C"    # Red - Has synovitis
)
```

---

# 4. Load Data

```{r load-data}
# =============================================================================
# PURPOSE: Load Olink NPX data and clinical metadata
#
# DATA SOURCES:
# - Olink: NPX values (normalized protein expression)
# - Metadata: Clinical variables including Joint_synovitis
# =============================================================================

# Load Olink data
olink <- read_parquet(here::here("data", "P2024-199-OLI_80_NPX_2024-10-10.parquet"))

# Load metadata (sheet 2 contains the arthritis cohort)
metadata <- read_excel(here::here("data", "Metadata_veSSC.xlsx"), sheet = 2)

cat("Olink samples:", length(unique(olink$SampleID)), "\n")
cat("Metadata samples:", nrow(metadata), "\n")
```

---

# 5. Data Preparation

```{r data-prep}
# =============================================================================
# PURPOSE: Prepare and merge data
#
# STEPS:
# 1. Create matching patient IDs
# 2. Filter for patients with Joint_synovitis data
# 3. Merge Olink with metadata
# =============================================================================

# Create matching IDs (trim whitespace, lowercase)
metadata$patient <- trimws(tolower(metadata$Barcode))
olink$patient <- trimws(tolower(olink$SampleID))

# Merge Olink with metadata
merged <- merge(olink, metadata, by = "patient", all.x = TRUE)

# Filter for patients with Joint_synovitis data (exclude controls)
filtered_df <- merged %>%
  filter(!is.na(Joint_synovitis))

# Ensure Joint_synovitis is a factor
filtered_df$Joint_synovitis <- factor(filtered_df$Joint_synovitis, 
                                       levels = c(0, 1), 
                                       labels = c("No", "Yes"))

cat("Patients with synovitis data:", length(unique(filtered_df$patient)), "\n")
cat("\nJoint synovitis breakdown:\n")
print(table(filtered_df$Joint_synovitis[!duplicated(filtered_df$patient)]))
```

---

# 6. Create Expression Matrix

```{r create-matrix}
# =============================================================================
# PURPOSE: Create protein expression matrix for limma
#
# WHY: limma needs a matrix with:
# - Rows = proteins (OlinkID)
# - Columns = samples (patients)
# - Values = NPX (expression)
# =============================================================================

# Create wide matrix
matrix_data <- filtered_df %>%
  dplyr::select(patient, OlinkID, NPX) %>%
  distinct() %>%
  pivot_wider(names_from = patient, values_from = NPX)

# Convert to matrix
data_matrix <- as.data.frame(matrix_data)
rownames(data_matrix) <- data_matrix$OlinkID
data_matrix <- data_matrix[, -1]  # Remove OlinkID column
data_matrix <- as.matrix(data_matrix)

# Convert to numeric
data_matrix2 <- apply(data_matrix, 2, as.numeric)
rownames(data_matrix2) <- rownames(data_matrix)

# Remove control samples (keep only patient samples)
keep_samples <- !grepl("neg|sample|plate|control", colnames(data_matrix2), ignore.case = TRUE)
data_matrix2 <- data_matrix2[, keep_samples]

cat("Expression matrix dimensions:", nrow(data_matrix2), "proteins x", ncol(data_matrix2), "samples\n")
```

---

# 7. Prepare Metadata for limma

```{r prep-metadata}
# =============================================================================
# PURPOSE: Prepare filtered metadata matching matrix columns
#
# CRITICAL: Metadata rows must match matrix columns EXACTLY
# =============================================================================

# Get unique metadata per patient
filtered_metadata <- metadata %>%
  filter(tolower(trimws(Barcode)) %in% colnames(data_matrix2)) %>%
  mutate(patient = tolower(trimws(Barcode)))

# Convert Immunosuppression to factor
filtered_metadata$Imunosuppression <- factor(filtered_metadata$Imunosuppression)

# Convert Joint_synovitis to factor
filtered_metadata$Joint_synovitis <- factor(filtered_metadata$Joint_synovitis,
                                             levels = c(0, 1),
                                             labels = c("No", "Yes"))

# =============================================================================
# EXTRACT VISIT YEAR FROM DATE
# =============================================================================

# Extract year from Visit_Date
filtered_metadata$Visit_Year <- factor(format(filtered_metadata$Visit_Date, "%Y"))

cat("Visit Year distribution:\n")
print(table(filtered_metadata$Visit_Year))

# Check for confounding with Joint_synovitis
cat("\nVisit Year by Joint Synovitis:\n")
print(table(filtered_metadata$Visit_Year, filtered_metadata$Joint_synovitis))

cat("Metadata prepared for", nrow(filtered_metadata), "samples\n")
cat("Joint synovitis:\n")
print(table(filtered_metadata$Joint_synovitis))
cat("\nImmunosuppression:\n")
print(table(filtered_metadata$Imunosuppression))
```

---
 
# 8. Sample Alignment (CRITICAL)

```{r alignment}
# =============================================================================
# PURPOSE: Ensure matrix columns and metadata rows are in the same order
#
# WHY THIS IS CRITICAL:
# limma matches by POSITION, not by name!
# If column 1 of matrix is patient A but row 1 of metadata is patient B,
# limma will assign patient B's group to patient A's expression data.
# =============================================================================

cat("=== BEFORE alignment ===\n")
cat("Matrix columns (first 5):", head(colnames(data_matrix2), 5), "\n")
cat("Metadata patients (first 5):", head(filtered_metadata$patient, 5), "\n")
cat("Aligned:", all(colnames(data_matrix2) == filtered_metadata$patient), "\n\n")

# Reorder metadata to match matrix columns
filtered_metadata <- filtered_metadata[match(colnames(data_matrix2), filtered_metadata$patient), ]

cat("=== AFTER alignment ===\n")
cat("Matrix columns (first 5):", head(colnames(data_matrix2), 5), "\n")
cat("Metadata patients (first 5):", head(filtered_metadata$patient, 5), "\n")
cat("Aligned:", all(colnames(data_matrix2) == filtered_metadata$patient), "\n")

# STOP if not aligned
stopifnot("Sample alignment failed!" = all(colnames(data_matrix2) == filtered_metadata$patient))
cat("\n✓ Alignment verified\n")
```

---

# 9. Binary Analysis: Joint Synovitis vs No Synovitis

```{r limma-analysis}
# =============================================================================
# PURPOSE: Test for differential protein abundance
#
# MODEL: ~0 + Joint_synovitis + Imunosuppression
# - ~0 + Joint_synovitis: estimate mean for each group (not intercept)
# - + Imunosuppression: adjust for immunosuppression status
#
# CONTRAST: Synovitis_Yes - Synovitis_No
# - Positive logFC = higher in patients WITH synovitis
# - Negative logFC = lower in patients WITH synovitis
# =============================================================================

# Create design matrix
design <- model.matrix(~0 + Joint_synovitis + Visit_Year, data = filtered_metadata)
colnames(design) <- gsub("Joint_synovitis", "", colnames(design))

cat("Design matrix:\n")
print(head(design))

# Fit model
fit <- lmFit(data_matrix2, design)

# Create contrast: Yes - No (positive = higher in synovitis)
contrast.matrix <- makeContrasts(
  Synovitis_Yes_vs_No = Yes - No,
  levels = design
)

# Apply contrast and empirical Bayes
fit2 <- contrasts.fit(fit, contrast.matrix)
fit2 <- eBayes(fit2)

# Extract results
results <- topTable(fit2, coef = 1, number = Inf, adjust.method = "BH")
results <- results %>% rownames_to_column(var = "OlinkID")

cat("\nTop 10 proteins by p-value:\n")
print(head(results, 10))
```

---

# 10. Expression Filtering & FDR Recalculation

```{r filter-fdr}
# =============================================================================
# PURPOSE: Filter for reliably detected proteins and recalculate FDR
#
# WHY FILTER ON AveExpr:
# - Low-expressed proteins are noisy and unreliable
# - AveExpr > 0 keeps proteins with detectable signal
#
# WHY RECALCULATE FDR:
# - Original adj.P.Val was calculated on ALL proteins
# - After filtering, fewer tests → less harsh correction
# - This is valid because AveExpr is INDEPENDENT of the test statistic
# =============================================================================

# Filter and recalculate FDR
results_filtered <- results %>%
  filter(AveExpr > 0) %>%
  mutate(adj.P.Val = p.adjust(P.Value, method = "BH")) %>%
  arrange(P.Value)

cat("Before filtering:", nrow(results), "proteins\n")
cat("After AveExpr > 0:", nrow(results_filtered), "proteins\n")
cat("FDR < 0.3:", sum(results_filtered$adj.P.Val < 0.2), "proteins\n")
```

---

# 11. Add Gene Names

```{r add-gene-names}
# =============================================================================
# PURPOSE: Add human-readable gene names to results
# =============================================================================

# Get unique OlinkID to Assay mapping
gene_map <- filtered_df %>%
  dplyr::select(OlinkID, Assay) %>%
  distinct()

# Merge with results
results_filtered <- results_filtered %>%
  left_join(gene_map, by = "OlinkID")

cat("Results with gene names:\n")
print(head(results_filtered[, c("OlinkID", "Assay", "logFC", "P.Value", "adj.P.Val")], 15))
```

---

# 12. Direction of Change Summary

```{r direction}
# =============================================================================
# PURPOSE: Summarize which proteins are up vs down in synovitis
#
# INTERPRETATION:
# - Positive logFC = higher in patients WITH joint synovitis
# - Negative logFC = lower in patients WITH joint synovitis
# =============================================================================

# Define hits
hits <- results_filtered %>% filter(adj.P.Val < 0.3)

hits_up <- hits %>% filter(logFC > 0)
hits_down <- hits %>% filter(logFC < 0)

cat("=== Direction of Change (FDR < 0.2) ===\n\n")
cat("Higher in Synovitis (positive logFC):", nrow(hits_up), "\n")
if(nrow(hits_up) > 0) cat("  ", paste(hits_up$Assay, collapse = ", "), "\n\n")

cat("Lower in Synovitis (negative logFC):", nrow(hits_down), "\n")
if(nrow(hits_down) > 0) cat("  ", paste(hits_down$Assay, collapse = ", "), "\n")
```

---

# 13. Volcano Plot

```{r volcano, fig.width=8, fig.height=7}
# =============================================================================
# PURPOSE: Visualize differential expression across all proteins
#
# X-axis: log2 fold change (effect size)
# Y-axis: -log10 p-value (significance)
# =============================================================================

# Add significance categories
results_filtered <- results_filtered %>%
  mutate(
    sig_category = case_when(
      adj.P.Val < 0.3 & logFC > 0.5 ~ "Up in Synovitis",
      adj.P.Val < 0.2 & logFC < -0.5 ~ "Down in Synovitis",
      TRUE ~ "NS"
    )
  )

# Select top hits to label
top_hits <- results_filtered %>%
  filter(adj.P.Val < 0.2) %>%
  arrange(P.Value) %>%
  head(15)

# Colors
colors <- c("Up in Synovitis" = "#E74C3C",
            "Down in Synovitis" = "#3498DB", 
            "NS" = "#CCCCCC")

# Create volcano plot
volcano <- ggplot(results_filtered, aes(x = logFC, y = -log10(P.Value))) +
  geom_point(aes(color = sig_category), size = 1.5, alpha = 0.6) +
  geom_text_repel(
    data = top_hits,
    aes(label = Assay),
    size = 3,
    max.overlaps = 20,
    segment.color = "gray50",
    segment.size = 0.3
  ) +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed", color = "gray40") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray40") +
  scale_color_manual(values = colors, name = "") +
  labs(
    title = "Joint Synovitis vs No Synovitis",
    subtitle = "Positive logFC = higher in synovitis patients",
    x = expression(Log[2]~Fold~Change),
    y = expression(-Log[10]~P-value)
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    plot.subtitle = element_text(hjust = 0.5, color = "gray40"),
    legend.position = "top"
  )

# Save
ggsave(file.path(output_dir, "volcano_synovitis.pdf"), volcano, width = 8, height = 7)

print(volcano)
```

---

# 14. Boxplots for Significant Proteins

```{r boxplots, fig.width=12, fig.height=15}
# =============================================================================
# PURPOSE: Visualize expression of top proteins
# =============================================================================

# Create plot data (wide format with metadata)
olink_wide <- filtered_df %>%
  dplyr::select(patient, Assay, NPX) %>%
  distinct() %>%
  pivot_wider(names_from = Assay, values_from = NPX)

# Merge with metadata
plot_data_all <- olink_wide %>%
  inner_join(filtered_metadata %>% dplyr::select(patient, Joint_synovitis), by = "patient")

# Select top 20 by p-value
top20 <- results_filtered %>%
  arrange(P.Value) %>%
  head(20)

# Plot function
plot_protein_violin <- function(data, protein_name, fdr_value) {
  
  ggplot(data, aes(x = Joint_synovitis, y = .data[[protein_name]], fill = Joint_synovitis)) +
    geom_violin(alpha = 0.5, color = NA, trim = FALSE) +
    geom_boxplot(width = 0.2, alpha = 0.9, outlier.shape = NA) +
    geom_jitter(width = 0.1, size = 1.5, alpha = 0.7, color = "black") +
    scale_fill_manual(values = colours_class) +
    stat_compare_means(method = "wilcox.test", label = "p.format", 
                       label.x.npc = 0.5, label.y.npc = 0.95, size = 3) +
    labs(
      title = protein_name,
      subtitle = paste0("FDR = ", round(fdr_value, 3)),
      x = NULL,
      y = "NPX"
    ) +
    theme_classic(base_size = 11) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5, color = "gray40"),
      legend.position = "none"
    )
}

# Generate plots
plot_list <- list()

for (i in 1:nrow(top20)) {
  protein <- top20$Assay[i]
  fdr <- top20$adj.P.Val[i]
  
  if (protein %in% colnames(plot_data_all)) {
    plot_list[[protein]] <- plot_protein_violin(plot_data_all, protein, fdr)
  }
}

# Combine plots
combined_plot <- plot_grid(plotlist = plot_list, ncol = 4, align = "hv")

# Save
ggsave(file.path(output_dir, "boxplots_synovitis_top20.pdf"), 
       combined_plot, width = 12, height = 15)

print(combined_plot)
```

---

# 15. Export Results

```{r export}
# =============================================================================
# PURPOSE: Save results for downstream analysis
# =============================================================================

# All results (filtered)
write.csv(results_filtered, 
          file.path(output_dir, "results_Synovitis_vs_NoSynovitis_all.csv"), 
          row.names = FALSE)

# Significant hits only
write.csv(hits, 
          file.path(output_dir, "results_Synovitis_vs_NoSynovitis_FDR02.csv"), 
          row.names = FALSE)

cat("Results exported to:", output_dir, "\n")
cat("Files:\n")
cat("  - results_Synovitis_vs_NoSynovitis_all.csv\n")
cat("  - results_Synovitis_vs_NoSynovitis_FDR02.csv\n")
cat("  - volcano_synovitis.pdf\n")
cat("  - boxplots_synovitis_top20.pdf\n")
```

---

# Session Info

```{r session-info}
sessionInfo()
```
