---
title: "Pathway Enrichment Analysis (Analysis D)"
author: "Analysis Script"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", dpi = 600)
suppressPackageStartupMessages({
  library(here); library(dplyr); library(ggplot2)
  library(clusterProfiler); library(org.Hs.eg.db)
  library(ReactomePA); library(enrichplot)
})

# All outputs live under output/<analysis_name>/ , resolved from the project
# root by the here package (robust to working directory).
analysis_name <- "RESOLVE_EnrichmentAnalysis"
output_dir    <- here::here("output", analysis_name)
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

# Color palettes
colours_eight <- c("#a4161a", "#F3722C", "#F9C74F", "#90BE6D", "#43AA8B", "#4D908E", "#3d348b", "#277DA1")
colors_class <- c("Regressor" = "#48639C", "Non_Regressor" = "#AF3B6E", "Stable" = "#ECA72C", "Healthy" = "#84A07C")
```

# Rationale

Over-representation analysis (ORA) asks whether the hit proteins fall into a
pathway more often than expected **by chance drawn from the same pool they came
from**. On a targeted Olink panel that pool is *not* the genome — the panel is
deliberately enriched for secreted, immune and inflammatory proteins. Testing
against a genome background would therefore credit the panel's own composition
as if it were biology. The correct background (`universe`) is the set of
proteins actually **tested** here: those passing the `AveExpr > 0` detection
filter, which is also the only pool from which a hit could be drawn.

GSEA needs no `universe`: it scores the full ranked list, so the ranking itself
is the background.

# Load results, define tested set and hits

```{r load-and-define}
results <- read.csv(here::here("output", "SSc_Skin_Regression_Analysis",
                               "results_continuous_Regressor_vs_Progressor.csv"))

# tested set: eligible-to-be-hit pool AND the ORA background
tested <- results %>%
  filter(AveExpr > 0) %>%
  mutate(adj.P.Val = p.adjust(P.Value, method = "BH"))

hits_fdr <- tested %>% filter(adj.P.Val < 0.2) %>% arrange(P.Value)

cat("Panel:", nrow(results),
    " | tested (AveExpr>0):", nrow(tested),
    " | hits (FDR<0.2):", nrow(hits_fdr), "\n")
cat("Up in Regressors:", sum(hits_fdr$logFC > 0),
    " | Up in Progressors:", sum(hits_fdr$logFC < 0), "\n")
```

# Map symbols to Entrez (hits and universe, identically)

```{r map-ids}
hit_entrez      <- bitr(hits_fdr$Assay, "SYMBOL", "ENTREZID", org.Hs.eg.db)
universe_entrez <- bitr(tested$Assay,   "SYMBOL", "ENTREZID", org.Hs.eg.db)

cat("Mapped hits:", nrow(hit_entrez), "of", nrow(hits_fdr),
    " | mapped universe:", nrow(universe_entrez), "of", nrow(tested), "\n")

hit_ids  <- hit_entrez$ENTREZID
univ_ids <- universe_entrez$ENTREZID
```

# GO over-representation (panel-aware background)

```{r go-ora}

run_go <- function(ont) {
  enrichGO(gene = hit_ids, universe = univ_ids, OrgDb = org.Hs.eg.db,
           ont = ont, pAdjustMethod = "BH",
           pvalueCutoff = 0.05, qvalueCutoff = 0.2, readable = TRUE)
}
go_bp <- run_go("BP"); go_mf <- run_go("MF"); go_cc <- run_go("CC")

# Collapse semantically redundant GO terms (keeps the most significant
# representative of each cluster). cutoff = similarity threshold (0.7 = default);
# lower it (e.g. 0.6) for more aggressive pruning.
simplify_go <- function(obj) {
  if (is.null(obj) || nrow(obj) == 0) return(obj)
  simplify(obj, cutoff = 0.7, by = "p.adjust", select_fun = min, measure = "Wang")
}
go_bp <- simplify_go(go_bp); go_mf <- simplify_go(go_mf); go_cc <- simplify_go(go_cc)

for (nm in c("BP","MF","CC")) {
  g <- get(paste0("go_", tolower(nm)))
  cat(sprintf("GO %s (after simplify): %d terms\n", nm, if (is.null(g)) 0 else nrow(g)))
}
```


## GO Biological Process

```{r go-bp-plot, fig.height = 10, fig.width =10}
if (!is.null(go_bp) && nrow(go_bp) > 0) {
  p <- dotplot(go_bp, showCategory = 15, title = "GO Biological Process")
  ggsave(here::here("output", analysis_name, "GO_BP_dotplot.pdf"), p, width = 10, height = 10)
  print(p)
} else cat("No significant GO-BP terms.\n")
```

## GO Molecular Function

```{r go-mf-plot, fig.height = 10, fig.width =10}
if (!is.null(go_mf) && nrow(go_mf) > 0) {
  p <- dotplot(go_mf, showCategory = 15, title = "GO Molecular Function")
  ggsave(here::here("output", analysis_name, "GO_MF_dotplot.pdf"), p, width = 10, height = 10)
  print(p)
} else cat("No significant GO-MF terms.\n")
```

## GO Cellular Component

```{r go-cc-plot, fig.height = 10, fig.width =10}
if (!is.null(go_cc) && nrow(go_cc) > 0) {
  p <- dotplot(go_cc, showCategory = 15, title = "GO Cellular Component")
  ggsave(here::here("output", analysis_name, "GO_CC_dotplot.pdf"), p, width = 10, height = 10)
  print(p)
} else cat("No significant GO-CC terms.\n")
```

# KEGG and Reactome (same background)

```{r kegg-reactome}
kegg <- enrichKEGG(gene = hit_ids, universe = univ_ids,
                   organism = "hsa", pvalueCutoff = 0.05, pAdjustMethod = "BH")
# enrichKEGG returns Entrez IDs in geneID -> convert to symbols for traceability
if (!is.null(kegg) && nrow(kegg) > 0)
  kegg <- setReadable(kegg, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")

reactome <- enrichPathway(gene = hit_ids, universe = univ_ids,
                          pvalueCutoff = 0.05, pAdjustMethod = "BH", readable = TRUE)

cat("KEGG:", if (is.null(kegg)) 0 else nrow(kegg), "terms | Reactome:",
    if (is.null(reactome)) 0 else nrow(reactome), "terms\n")
```

## KEGG

```{r kegg-plot, fig.height = 10, fig.width =10}
if (!is.null(kegg) && nrow(kegg) > 0) {
  p <- dotplot(kegg, showCategory = 15, title = "KEGG Pathways")
  ggsave(here::here("output", analysis_name, "KEGG_dotplot.pdf"), p, width = 10, height = 10)
  print(p)
} else cat("No significant KEGG terms.\n")
```

## Reactome

```{r reactome-plot, fig.height = 10, fig.width =10}
if (!is.null(reactome) && nrow(reactome) > 0) {
  p <- dotplot(reactome, showCategory = 15, title = "Reactome Pathways")
  ggsave(here::here("output", analysis_name, "Reactome_dotplot.pdf"), p, width = 10, height = 8)
  print(p)
} else cat("No significant Reactome terms.\n")
```

# GSEA (full ranked tested list, no universe)

```{r gsea}
ranks <- tested %>%
  inner_join(bitr(tested$Assay, "SYMBOL", "ENTREZID", org.Hs.eg.db),
             by = c("Assay" = "SYMBOL")) %>%
  arrange(desc(t))
gene_list <- setNames(ranks$t, ranks$ENTREZID)

set.seed(1)
gsea_go <- gseGO(gene_list, OrgDb = org.Hs.eg.db, ont = "BP",
                 pAdjustMethod = "BH", pvalueCutoff = 0.05, verbose = FALSE)

# simplify works on gseGO results too; then convert leading-edge IDs to symbols
if (!is.null(gsea_go) && nrow(gsea_go) > 0) {
  gsea_go <- simplify(gsea_go, cutoff = 0.7, by = "p.adjust", select_fun = min)
  gsea_go <- setReadable(gsea_go, OrgDb = org.Hs.eg.db, keyType = "ENTREZID")
}
cat("GSEA GO-BP (after simplify):", if (is.null(gsea_go)) 0 else nrow(gsea_go), "terms\n")
```

```{r gsea-plot, fig.height = 12, fig.width=12}
if (!is.null(gsea_go) && nrow(gsea_go) > 0) {
  p <- dotplot(gsea_go, showCategory = 10, split = ".sign",
               title = "GSEA GO-BP (NES)") + facet_grid(. ~ .sign)
  ggsave(here::here("output", analysis_name, "GSEA_GO_BP_dotplot.pdf"), p, width = 11, height = 8)
  print(p)
} else cat("No significant GSEA GO-BP terms.\n")
```

# Result tables

```{r write-tables}
wr <- function(obj, file) if (!is.null(obj) && nrow(obj) > 0)
  write.csv(as.data.frame(obj), here::here("output", analysis_name, file), row.names = FALSE)
wr(go_bp, "GO_BP.csv"); wr(go_mf, "GO_MF.csv"); wr(go_cc, "GO_CC.csv")
wr(kegg, "KEGG.csv"); wr(reactome, "Reactome.csv")
if (!is.null(gsea_go)) wr(gsea_go, "GSEA_GO_BP.csv")

if (!is.null(go_bp) && nrow(go_bp) > 0)
  knitr::kable(head(as.data.frame(go_bp)[, c("Description","GeneRatio","BgRatio","p.adjust")], 15),
               caption = "Top GO-BP terms (panel-aware background)")
```

---

# Session Info

```{r session-info}
sessionInfo()
```
