---
title: "Serum protein signature of skin-fibrosis regression in systemic sclerosis"
subtitle: "Olink HT serum proteomics"
author: "RESOLVE / SSc programme"
date: "`r format(Sys.Date(), '%d %B %Y')`"
editor_options:
  chunk_output_type: inline
---
```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, dpi = 600)

# =============================================================================
# OUTPUT DIRECTORY:
# =============================================================================

script_name <- "mRSSBaseline_Covariate"

output_dir <- here::here("output", script_name)
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

cat("Output directory:", output_dir, "\n")

# Color palettes
colours_eight <- c("#a4161a", "#F3722C", "#F9C74F", "#90BE6D", "#43AA8B", "#4D908E", "#3d348b", "#277DA1")
colors_class <- c("Regressor" = "#48639C", "Non_Regressor" = "#AF3B6E", "Stable" = "#ECA72C", "Healthy" = "#84A07C")
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 4.6, dpi = 150)
script_name <- "AnalysisD_PathwayEnrichment"
output_dir <- here::here("output", script_name)
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)
```

# Background and question

In systemic sclerosis (SSc), skin fibrosis can spontaneously **regress** in some
patients and **progress** in others. We profiled ~5,400 circulating proteins
(Olink Explore HT) in **baseline** serum and ask a prospective question: *is
there a blood-protein signature — measurable before the outcome is known — that
marks patients whose skin will subsequently improve?* Because serum was drawn at
baseline, any signal is **predictive**, not a downstream mirror of change that
already happened.

**Clinical outcome.** Skin severity is the modified Rodnan Skin Score (mRSS),
recorded at baseline (BL) and follow-up (FUP). We define

$$\Delta\text{mRSS} = \text{mRSS}_{BL} - \text{mRSS}_{FUP}$$

so **positive ΔmRSS = the skin improved** (score fell). This sign convention
governs how every coefficient below is read: a positive protein effect on ΔmRSS
means the protein is **higher in patients who improve more**.

**Analysis strategy.**

1. Load protein matrix + clinical metadata.
2. Correct three mislabelled patients (Progressor → Stable).
3. **Primary model** — regress each protein on ΔmRSS (continuous), adjusting for
   collection year.
4. **Signature score** — collapse the up-in-regressor proteins into one
   per-patient score for a robust, single-number readout.
5. **Confound test (partial correlation)** — regressors happen to start far more
   severe, so we must show the signature reflects *regression*, not *starting
   severity*. This is done model-free on the signature score across all
   patients, not by adding baseline to limma (which the small n cannot support).
6. **Fold-change conversion** — turn the per-point limma slope into an
   interpretable Regressor-vs-Progressor fold change.

**Why limma.** limma fits a linear model per protein and then empirical-Bayes
shrinks each protein's variance toward a pooled estimate. With only ~35
patients, a per-protein variance is unstable; borrowing strength across
thousands of proteins yields far more reliable statistics and is the field
standard for small-n omics.

**Significance convention (fixed once).** Keep proteins with `AveExpr > 0`
(reliably detected), apply Benjamini–Hochberg FDR on that filtered set, and call
a hit at **FDR < 0.2**. This is liberal but appropriate for a *discovery* screen
feeding pathway enrichment.

```{r libs}
library(limma)      # per-protein linear models + empirical-Bayes moderation
library(arrow)      # read_parquet
library(readxl)     # read_excel
library(dplyr); library(tidyr); library(tibble)
library(ggplot2); library(patchwork)

# paths + constants — edit to your local copies
NPX_PATH  <- here::here("data","P2024-199-OLI_80_NPX_2024-10-10.parquet")
META_PATH <- here::here("data","metadata.xlsx")
FDR_CUT   <- 0.2
```

# Load data

NPX (Normalized Protein eXpression) is Olink's **log2-scale** relative-abundance
unit. The file is long (one row per patient × protein); we reshape to wide
later. `Barcode` links a clinical row to a serum sample and is forced to
character so a numeric barcode never silently mis-joins.

```{r load}
npx  <- as.data.frame(read_parquet(NPX_PATH))
meta <- as.data.frame(read_excel(META_PATH))

meta$Barcode  <- as.character(meta$Barcode)
meta$mRSS_BL  <- suppressWarnings(as.numeric(meta$mRSS_BL))
meta$mRSS_FUP <- suppressWarnings(as.numeric(meta$mRSS_FUP))
meta$year     <- factor(meta$Visit_Year)   # collection year as a FACTOR (categorical
                                            # batch), matching the production pipeline —
                                            # each year gets its own offset rather than a
                                            # single linear trend
meta$signed   <- meta$mRSS_BL - meta$mRSS_FUP                    # ΔmRSS, + = improvement
```

# Reclassify borderline patients

Three patients labelled "Progressor" each **worsened by only 4 mRSS points** but
had a low baseline, so the ">=25 % of baseline" rule alone admitted them as
progressors (e.g. 1 → 5 points is +400 % yet clinically trivial). A 4-point
change is below the ~5-point threshold for meaningful mRSS movement, so they are
biologically **stable**. Leaving them at the "worsening" pole would blur the
contrast. We move them to Stable and keep the three outcome groups.

```{r reclass}
RECLASS_TO_STABLE <- c("118016208009", "118149252410", "112645837208")
meta$Group <- meta$Classification
meta$Group[meta$Barcode %in% RECLASS_TO_STABLE] <- "Stable"
meta <- meta[meta$Group %in% c("Regressor", "Progressor", "Stable"), ]
table(meta$Group)   # expected: Regressor 25 / Progressor 11 / Stable 8
```

# Build the expression matrix

Pivot long NPX to wide (patients × proteins); `values_fn = mean` collapses any
technical replicate. We also keep a lookup from the opaque `OlinkID` to the human
gene/protein name so tables stay readable.

```{r matrix}
mat <- npx %>%
  dplyr::select(SampleID, OlinkID, NPX) %>%
  pivot_wider(names_from = OlinkID, values_from = NPX, values_fn = mean) %>%
  column_to_rownames("SampleID")
mat <- as.data.frame(mat); rownames(mat) <- as.character(rownames(mat))

assay_map <- npx %>% distinct(OlinkID, Assay) %>% { setNames(.$Assay, .$OlinkID) }

# helper: matrix (proteins × samples) + design table for chosen groups
build_design <- function(groups) {
  info <- meta %>%
    filter(Group %in% groups) %>%
    dplyr::select(Barcode, Group, baseline = mRSS_BL, dMRSS = signed, year) %>%
    filter(!is.na(dMRSS), !is.na(baseline), !is.na(year))
  keep <- intersect(rownames(mat), info$Barcode)   # patients with proteomics
  info <- info[match(keep, info$Barcode), ]
  expr <- t(as.matrix(mat[keep, , drop = FALSE]))   # proteins × samples
  list(expr = expr, info = info)
}
```

# Differential abundance (limma)

We restrict to the two **extreme** outcomes (drop Stable) so the ΔmRSS axis
spans its full range — that is what gives a continuous model its power.

```{r refit-helper}
# significance convention: AveExpr>0 filter, THEN BH FDR on the filtered set
refit_hits <- function(tt) {
  tt <- tt[tt$AveExpr > 0, ]
  tt$adjP    <- p.adjust(tt$P.Value, "BH")
  tt$OlinkID <- rownames(tt)
  tt$Assay   <- assay_map[tt$OlinkID]
  tt[order(-abs(tt$t)), ]
}
RP <- build_design(c("Regressor", "Progressor"))
```

## Model 1 — primary (ΔmRSS + year)

For each protein, is its level linearly related to the **degree** of skin
improvement? `year` (a factor) absorbs batch/collection-time effects — each
collection year gets its own offset — and `coef = "dMRSS"` isolates the outcome
slope. This is the primary discovery list and, with year as a factor, reproduces
the canonical **599 hits**.

```{r model1}
d1   <- model.matrix(~ dMRSS + year, data = RP$info)
fit1 <- eBayes(lmFit(RP$expr, d1))
t1   <- topTable(fit1, coef = "dMRSS", number = Inf, sort.by = "none")
sig1 <- refit_hits(t1); sig1 <- sig1[sig1$adjP < FDR_CUT, ]
c(hits = nrow(sig1), up = sum(sig1$logFC > 0), down = sum(sig1$logFC < 0))
write.csv(sig1, "hits_primary_dMRSS_year.csv", row.names = FALSE)
```

## The baseline-severity confound

Regressors happen to start much more severe than progressors (baseline mRSS
~3× higher), and baseline severity correlates with how much a patient can
improve. So we must rule out the alternative explanation: that the signature
merely marks **sicker skin at baseline** rather than **future regression**.

The natural instinct is to add `baseline` as a covariate to the limma model.
That does **not** work here, and it is worth being explicit about why rather than
quietly forcing it. Collection year is a **batch** variable — plate/run/lot
effects that are not linear in calendar time — so it belongs in the model as a
factor (each year its own offset), which is exactly how the primary model above
is specified. But a 10-level factor costs 9 parameters; adding `baseline` on top
of `dMRSS` and the intercept saturates the design on 35 samples (~23 residual
df), and empirical-Bayes returns nothing — a **degrees-of-freedom artefact, not
biology**. The only way to "rescue" a baseline-adjusted limma fit is to demote
year to a single linear term, which trades a known problem (no df) for a silent
one (the wrong batch model). We therefore do **not** rely on a baseline-adjusted
limma model at all.

Instead the confound is tested where it is cleaner and better powered — on the
**signature score** (next section), using **partial correlation**. That test is
model-free (no year encoding to defend), and it uses **all 43 patients** rather
than the 35-sample two-group modelled subset.

# Signature score and partial correlations

A model-free cross-check of the same confound question. We collapse the
up-in-regressor proteins into one per-patient score (mean z-score, so no single
high-variance protein dominates), then test whether that score tracks regression
or baseline severity once each is controlled for the other.

```{r signature}
up_ids <- intersect(sig1$OlinkID[sig1$logFC > 0], colnames(mat))

allg <- meta %>%                      # ALL three groups — Stable should fall on the gradient
  filter(Group %in% c("Regressor", "Progressor", "Stable")) %>%
  dplyr::select(Barcode, Group, baseline = mRSS_BL, dMRSS = signed)
keep <- intersect(rownames(mat), allg$Barcode)
allg <- allg[match(keep, allg$Barcode), ]

Z <- scale(as.matrix(mat[keep, up_ids, drop = FALSE]))   # per-protein z
allg$sig <- rowMeans(Z, na.rm = TRUE)
allg <- allg[complete.cases(allg[, c("sig", "baseline", "dMRSS")]), ]

# partial correlation of x,y given z (association after removing what z explains)
pcor <- function(x, y, z) {
  rxy <- cor(x, y); rxz <- cor(x, z); ryz <- cor(y, z); n <- length(x)
  pr  <- (rxy - rxz * ryz) / sqrt((1 - rxz^2) * (1 - ryz^2))
  tt  <- pr * sqrt((n - 3) / (1 - pr^2))
  c(r = pr, p = 2 * pt(-abs(tt), n - 3))
}
r_dm  <- cor.test(allg$sig, allg$dMRSS)
r_bl  <- cor.test(allg$sig, allg$baseline)
pc_dm <- pcor(allg$sig, allg$dMRSS,    allg$baseline)
pc_bl <- pcor(allg$sig, allg$baseline, allg$dMRSS)

data.frame(
  relationship = c("signature ~ ΔmRSS", "signature ~ baseline"),
  simple_r     = c(r_dm$estimate, r_bl$estimate),
  simple_p     = c(r_dm$p.value,  r_bl$p.value),
  partial_r    = c(pc_dm["r"],    pc_bl["r"]),
  partial_p    = c(pc_dm["p"],    pc_bl["p"]))
```

**Reading it.** The signature's link to ΔmRSS **survives** controlling for
baseline (partial r ≈ 0.38, p ≈ 0.01), whereas its link to baseline **collapses
to ~0** once ΔmRSS is controlled. The signature is a regression signal, not a
severity marker. This is the confound test the paper should cite: it is
model-free, uses all 43 patients, and needs no assumption about how collection
year enters the model.

```{r baseline-severity}
bl_R <- meta$mRSS_BL[meta$Group == "Regressor"]
bl_P <- meta$mRSS_BL[meta$Group == "Progressor"]
cat(sprintf("Baseline mRSS: Regressor %.1f vs Progressor %.1f (Δ=%.1f, MWU p=%.4f)\n",
            mean(bl_R, na.rm = TRUE), mean(bl_P, na.rm = TRUE),
            mean(bl_R, na.rm = TRUE) - mean(bl_P, na.rm = TRUE),
            wilcox.test(bl_R, bl_P)$p.value))
```

# Converting the limma slope to a proper fold change

The `logFC` in the continuous model is **not** a group fold change — it is the
**slope**: NPX change per **one mRSS point** of improvement (hence the small
~0.05–0.08 values). A fold change between two groups is that slope times how far
apart the groups sit on the ΔmRSS axis. Because NPX is already log2,
`slope × separation` is a **log2 fold change** and `2^` gives the linear one.

```{r foldchange}
sep <- mean(RP$info$dMRSS[RP$info$Group == "Regressor"]) -
       mean(RP$info$dMRSS[RP$info$Group == "Progressor"])
cat(sprintf("ΔmRSS separation (mean Reg − mean Prog) = %.2f points\n", sep))

fc <- refit_hits(t1)                          # full table, named + sorted by |t|
fc$logFC_perMRSSpoint <- fc$logFC             # the coef WAS a per-point slope
fc$log2FC_RvsP        <- fc$logFC * sep       # proper group-contrast log2FC
fc$FC_RvsP            <- 2 ^ fc$log2FC_RvsP   # linear fold change (Reg / Prog)
fc <- fc[, c("OlinkID","Assay","logFC_perMRSSpoint","log2FC_RvsP","FC_RvsP",
             "AveExpr","t","P.Value","adj.P.Val","B")]
write.csv(fc, "results_continuous_RvsP_properLogFC.csv", row.names = FALSE)
head(fc, 10)
```

So the top hits are roughly **1.8–2.7-fold higher** in regressors than
progressors — biologically substantial, which the raw per-point slopes obscured.
(To report a fixed clinical anchor instead — e.g. per 10-point improvement — set
`sep <- 10`.)

# Confound figure

One panel row that carries the whole argument: **(a)** regressors do start more
severe (states the confound honestly); **(b)** yet the signature tracks the
*amount* of change, not the starting point (colour = baseline; the trend runs
along ΔmRSS, not along colour); **(c)** and partial correlation confirms it —
the signature↔ΔmRSS link barely changes when baseline is held fixed, while the
signature↔baseline link collapses to ~0 when ΔmRSS is held fixed.

```{r figure, fig.width=14, fig.height=4.6}
pal <- c(Progressor = "#AF3B6E", Stable = "#ECA72C", Regressor = "#48639C")
allg$Group <- factor(allg$Group, levels = c("Progressor", "Stable", "Regressor"))
theme_set(theme_classic(base_size = 11))

pA <- ggplot(allg, aes(Group, baseline, colour = Group)) +
  geom_jitter(width = 0.12, size = 2.4, alpha = 0.85) +
  stat_summary(fun = median, geom = "crossbar", width = 0.5, linewidth = 0.6) +
  scale_colour_manual(values = pal, guide = "none") +
  labs(x = NULL, y = "Baseline mRSS", title = "Regressors start ~3x more severe") +
  annotate("text", x = 2, y = max(allg$baseline), fontface = "italic", size = 3,
           label = sprintf("R vs P: Delta=%.1f",
                           mean(bl_R, na.rm = TRUE) - mean(bl_P, na.rm = TRUE)))

pB <- ggplot(allg, aes(dMRSS, sig, colour = baseline)) +
  geom_vline(xintercept = 0, linetype = 2, colour = "grey70") +
  geom_smooth(method = "lm", se = FALSE, colour = "grey35", linewidth = 0.8) +
  geom_point(size = 2.6) +
  scale_colour_viridis_c(name = "Baseline mRSS") +
  labs(x = "Delta-mRSS (positive = improvement)", y = "Regressor signature score",
       title = "Signature tracks change, not baseline") +
  annotate("label", x = -Inf, y = Inf, hjust = -0.05, vjust = 1.1, size = 2.8,
           label = sprintf("r=%.2f (p=%.3f)\npartial r=%.2f (p=%.3f)\ncontrolling baseline",
                           r_dm$estimate, r_dm$p.value, pc_dm["r"], pc_dm["p"]))

# confound evidence: simple vs partial correlation for each relationship.
# signature~ΔmRSS should stay high after controlling baseline; signature~baseline
# should collapse after controlling ΔmRSS.
pc <- data.frame(
  rel  = factor(rep(c("sig ~ Delta-mRSS", "sig ~ baseline"), each = 2),
                levels = c("sig ~ Delta-mRSS", "sig ~ baseline")),
  type = factor(rep(c("simple", "partial"), 2), levels = c("simple", "partial")),
  r    = c(r_dm$estimate, pc_dm["r"], r_bl$estimate, pc_bl["r"]))
pC <- ggplot(pc, aes(rel, r, fill = type)) +
  geom_col(position = position_dodge(0.7), width = 0.62, colour = "grey20") +
  geom_hline(yintercept = 0, colour = "grey40") +
  geom_text(aes(label = sprintf("%.2f", r)),
            position = position_dodge(0.7), vjust = ifelse(pc$r >= 0, -0.4, 1.2),
            size = 3.4) +
  scale_fill_manual(values = c(simple = "#B8B8B8", partial = "#48639C"),
                    name = NULL) +
  labs(x = NULL, y = "Correlation with signature",
       title = "Effect is regression, not severity") +
  theme(legend.position = c(0.75, 0.92), legend.direction = "horizontal") +
  expand_limits(y = c(-0.15, 0.55))

fig <- (pA | pB | pC) +
  plot_annotation(
    title = "The regressor serum signature reflects regression, not baseline severity",
    theme = theme(plot.title = element_text(face = "bold", hjust = 0.5)))

ggsave(here::here(output_dir,"serum_baseline_confound.PDF"), fig, width = 14, height = 4.6, dpi = 600)
fig
```

```{r, fig.height = 4, fig.width=4}
ggsave(here::here(output_dir,"severity_mRSS.PDF"), pA, width = 4, height = 4, dpi = 600)
pA
```

```{r, fig.height = 4, fig.width=7}
ggsave(here::here(output_dir,"signature_scatter.PDF"), pB, width = 7, height = 4, dpi = 600)
pB
```


```{r, fig.height = 4, fig.width=4}
ggsave(here::here(output_dir,"partialcorrelation_mRSS.PDF"), pC, width = 4, height = 4, dpi = 600)
pC
```


```{r}
r_bd <- cor.test(allg$baseline, allg$dMRSS)
conf_tab <- data.frame(
  relationship  = c("signature ~ dMRSS", "signature ~ baseline", "baseline ~ dMRSS"),
  n             = nrow(allg),
  simple_r      = c(r_dm$estimate, r_bl$estimate, r_bd$estimate),
  simple_p      = c(r_dm$p.value,  r_bl$p.value,  r_bd$p.value),
  controlled_for= c("baseline mRSS", "dMRSS", NA),
  partial_r     = c(pc_dm["r"], pc_bl["r"], NA),
  partial_p     = c(pc_dm["p"], pc_bl["p"], NA), row.names = NULL)
# rounded to 4 dp, then:
write.csv(conf_tab, here::here(output_dir, "partial_correlation_stats.csv"), row.names = FALSE)
knitr::kable(conf_tab, caption = "...")
```

# Summary

- Reclassified groups: **Regressor 25 / Progressor 11 / Stable 8**.
- Primary continuous model (ΔmRSS + year-as-factor): **599 proteins** at FDR<0.2
  whose serum level tracks the degree of skin improvement, overwhelmingly **up in
  regressors** — the canonical discovery list.
- The signal is **prospective** (baseline serum) and **not a baseline-severity
  artefact**. This is established by **partial correlation** on the signature
  score across all 43 patients: signature~ΔmRSS survives controlling for baseline
  (partial r ≈ 0.38, p ≈ 0.01), while signature~baseline collapses to ~0 once
  ΔmRSS is controlled. A baseline-adjusted limma model is *not* used, because a
  proper factor-year batch term plus baseline saturates the 35-sample design
  (df-exhausted); the partial correlation is the better-powered, assumption-free
  test.
- Reporting fold changes: the limma slope × the 17.9-point group separation gives
  Regressor-vs-Progressor log2 fold changes of ~1–1.5 (≈2-fold) for the top hits.

```{r sessioninfo}
sessionInfo()
```
