---
title: "Pathway Enrichment Analysis (Analysis D)"
author: "Analysis Script"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, dpi = 300)

# =============================================================================
# OUTPUT DIRECTORY: output/<script_name>/
# =============================================================================

script_name <- "AnalysisD_PathwayEnrichment"

output_dir <- here::here("output", script_name)
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

cat("Output directory:", output_dir, "\n")

# Color palettes
colours_eight <- c("#a4161a", "#F3722C", "#F9C74F", "#90BE6D", "#43AA8B", "#4D908E", "#3d348b", "#277DA1")
colors_class <- c("Regressor" = "#48639C", "Non_Regressor" = "#AF3B6E", "Stable" = "#ECA72C", "Healthy" = "#84A07C")
```

## Overview

Pathway enrichment analysis on protein hits from the **Regressor vs Progressor continuous analysis**.

Two approaches for defining hits:
1. **FDR < 0.2** — statistical threshold
2. **Top 20 proteins** — rank-based

```{r libraries}
# =============================================================================
# REQUIRED PACKAGES
# Install if needed:
#   BiocManager::install(c("clusterProfiler", "org.Hs.eg.db", "enrichplot", "ReactomePA"))
#   install.packages("msigdbr")
# =============================================================================

library(dplyr)
library(ggplot2)
library(tidyr)
library(here)
library(arrow)
library(readxl)
library(ggpubr)
library(cowplot)

# Enrichment analysis
library(clusterProfiler)    # GO, KEGG enrichment
library(org.Hs.eg.db)       # Human gene annotation
library(enrichplot)         # Visualization
library(ReactomePA)         # Reactome pathway analysis

# Gene sets
library(msigdbr)            # MSigDB gene sets (Hallmark, etc.)
```

---

# 1. Load Results

```{r load-results}
# =============================================================================
# PURPOSE: Load results from the continuous R vs P analysis
#
# This should be the output from your main analysis script
# =============================================================================

# Load the continuous R vs P results
results <- read.csv(here::here("output", "SSc_Skin_Regression_Analysis", "results_continuous_Regressor_vs_Progressor.csv"))

cat("Loaded", nrow(results), "proteins\n")
head(results)
```

### 1.1 Checking the distribution of average expression across the cohort

```{r aveexpr-hist}
summary(results$AveExpr)

hist_plot <- ggplot(data.frame(AveExpr = results$AveExpr), aes(x = AveExpr)) +
  geom_histogram(bins = 30, fill = "#3d348b", color = "white", alpha = 0.7) +
  geom_vline(xintercept = c(0, 1), linetype = "dashed", color = c("pink", "lightblue")) +
  labs(title = "AveExpr Distribution", x = "Average Expression", y = "Count") +
  theme_minimal()

ggsave(file.path(output_dir, "AveExpr_histogram.pdf"), hist_plot, width = 8, height = 6)
print(hist_plot)
```

---

# 2. Define Protein Hits

```{r define-hits}
# =============================================================================
# PURPOSE: Define significant proteins using two approaches
#
# 1. FDR < 0.2: Statistical threshold (more proteins for enrichment)
# 2. Top 20: Rank-based (ensures sufficient proteins for analysis)
#
# WHY TWO APPROACHES:
# - If FDR threshold yields few proteins, enrichment may fail
# - Top-N approach guarantees enough proteins for pathway analysis
# - Comparing both shows robustness of enrichment results
# =============================================================================

# Approach 1: FDR threshold
hits_fdr <- results %>%
  filter(AveExpr > 0) %>%
  mutate(adj.P.Val = p.adjust(P.Value, method = "BH")) %>%  # Recalculate!
  filter(adj.P.Val < 0.2) %>%
  arrange(P.Value)

cat("=== FDR < 0.2 ===\n")
cat("N proteins:", nrow(hits_fdr), "\n")
cat("Proteins:", paste(hits_fdr$Assay, collapse = ", "), "\n\n")

# Approach 2: Top 20 by p-value
hits_top20 <- results %>%
  filter(AveExpr > 0) %>%
  mutate(adj.P.Val = p.adjust(P.Value, method = "BH")) %>%
  arrange(P.Value) %>%
  head(20)

cat("=== Top 20 ===\n")
cat("Proteins:", paste(hits_top20$Assay, collapse = ", "), "\n")
cat("P-value range:", min(hits_top20$P.Value), "-", max(hits_top20$P.Value), "\n")
```

```{r direction-of-change}
# =============================================================================
# PURPOSE: Separate up- and down-regulated proteins
#
# WHY: Proteins higher in Regressors vs Progressors may reflect different
#      biology than proteins lower in Regressors
#
# INTERPRETATION (from continuous analysis):
# - Positive logFC: higher expression associated with MORE improvement
# - Negative logFC: higher expression associated with LESS improvement (worsening)
# =============================================================================

# Using FDR < 0.2 hits for this analysis
hits_up <- hits_fdr %>% filter(logFC > 0)
hits_down <- hits_fdr %>% filter(logFC < 0)

cat("=== Direction of Change (FDR < 0.2) ===\n")
cat("Higher in Regressors (positive logFC):", nrow(hits_up), "\n")
cat("  ", paste(hits_up$Assay, collapse = ", "), "\n\n")
cat("Higher in Progressors (negative logFC):", nrow(hits_down), "\n")
cat("  ", paste(hits_down$Assay, collapse = ", "), "\n")
```

---

# 3. Gene ID Conversion

```{r convert-ids}
# =============================================================================
# PURPOSE: Convert gene symbols to Entrez IDs
#
# WHY: clusterProfiler and most pathway databases use Entrez IDs
#      Gene symbols (like "MMP1") must be converted
#
# NOTE: Some genes may not map — this is normal (pseudogenes, novel genes)
# =============================================================================

# Convert gene symbols to Entrez IDs (using FDR < 0.2 hits)
gene_symbols <- hits_fdr$Assay

entrez_ids <- bitr(gene_symbols, 
                   fromType = "SYMBOL", 
                   toType = "ENTREZID", 
                   OrgDb = org.Hs.eg.db)

cat("Mapped", nrow(entrez_ids), "of", length(gene_symbols), "genes to Entrez IDs\n")

# Check which genes failed to map
unmapped <- setdiff(gene_symbols, entrez_ids$SYMBOL)
if (length(unmapped) > 0) {
  cat("Unmapped genes:", paste(unmapped, collapse = ", "), "\n")
}

# Create mapping for all results (for GSEA later)
all_entrez <- bitr(results$Assay, 
                   fromType = "SYMBOL", 
                   toType = "ENTREZID", 
                   OrgDb = org.Hs.eg.db)

cat("\nTotal mapped for GSEA:", nrow(all_entrez), "of", nrow(results), "\n")
```

---

# 4. GO Enrichment Analysis

```{r go-enrichment}
# =============================================================================
# PURPOSE: Gene Ontology enrichment analysis
#
# THREE ONTOLOGIES:
# - BP (Biological Process): What the proteins do together
# - MF (Molecular Function): Biochemical activities
# - CC (Cellular Component): Where proteins localize
#
# METHOD: Over-representation analysis (ORA)
# Tests if hit genes are enriched in GO terms vs background
# =============================================================================

# GO Biological Process
go_bp <- enrichGO(gene = entrez_ids$ENTREZID,
                  OrgDb = org.Hs.eg.db,
                  ont = "BP",
                  pAdjustMethod = "BH",
                  pvalueCutoff = 0.05,
                  qvalueCutoff = 0.2,
                  readable = TRUE)  # Convert back to gene symbols

cat("=== GO Biological Process ===\n")
if (!is.null(go_bp) && nrow(go_bp) > 0) {
  cat("Significant terms:", nrow(go_bp), "\n")
  print(head(go_bp@result[, c("Description", "GeneRatio", "pvalue", "p.adjust", "geneID")], 10))
} else {
  cat("No significant terms found\n")
}
```

```{r go-mf-cc}
# GO Molecular Function
go_mf <- enrichGO(gene = entrez_ids$ENTREZID,
                  OrgDb = org.Hs.eg.db,
                  ont = "MF",
                  pAdjustMethod = "BH",
                  pvalueCutoff = 0.05,
                  qvalueCutoff = 0.2,
                  readable = TRUE)

cat("\n=== GO Molecular Function ===\n")
if (!is.null(go_mf) && nrow(go_mf) > 0) {
  cat("Significant terms:", nrow(go_mf), "\n")
  print(head(go_mf@result[, c("Description", "GeneRatio", "pvalue", "p.adjust")], 10))
} else {
  cat("No significant terms found\n")
}

# GO Cellular Component
go_cc <- enrichGO(gene = entrez_ids$ENTREZID,
                  OrgDb = org.Hs.eg.db,
                  ont = "CC",
                  pAdjustMethod = "BH",
                  pvalueCutoff = 0.05,
                  qvalueCutoff = 0.2,
                  readable = TRUE)

cat("\n=== GO Cellular Component ===\n")
if (!is.null(go_cc) && nrow(go_cc) > 0) {
  cat("Significant terms:", nrow(go_cc), "\n")
  print(head(go_cc@result[, c("Description", "GeneRatio", "pvalue", "p.adjust")], 10))
} else {
  cat("No significant terms found\n")
}
```

```{r go-bp-plot, fig.width=10, fig.height=8}
if (!is.null(go_bp) && nrow(go_bp) > 0) {
  p1 <- dotplot(go_bp, showCategory = 15, title = "GO Biological Process")
  ggsave(file.path(output_dir, "GO_BP_dotplot.pdf"), p1, width = 10, height = 8)
  print(p1)
}
```

```{r go-mf-plot, fig.width=10, fig.height=8}
if (!is.null(go_mf) && nrow(go_mf) > 0) {
  p1 <- dotplot(go_mf, showCategory = 15, title = "GO Molecular Function")
  ggsave(file.path(output_dir, "GO_MF_dotplot.pdf"), p1, width = 10, height = 8)
  print(p1)
}
```

```{r go-cc-plot, fig.width=10, fig.height=8}
if (!is.null(go_cc) && nrow(go_cc) > 0) {
  p1 <- dotplot(go_cc, showCategory = 15, title = "GO Cellular Component")
  ggsave(file.path(output_dir, "GO_CC_dotplot.pdf"), p1, width = 10, height = 8)
  print(p1)
}
```

---

# 5. Boxplots for Hit Proteins

```{r load-expression-data, fig.height=15}
# Load Olink data
olink <- read_parquet(here::here("data", "P2024-199-OLI_80_NPX_2024-10-10.parquet"))

# Load metadata
metadata <- read_excel(here::here("data", "metadata.xlsx"))

# Prepare metadata
metadata <- metadata %>%
  filter(!EUSTAR_Number %in% c(180, 216)) %>%
  mutate(
    Classification = case_when(
      EUSTAR_Number %in% c(paste0("HC-OL-0", 1:5)) ~ "Healthy",
      TRUE ~ Classification
    ),
    EUSTAR_Number = as.character(EUSTAR_Number)
  )

# Reshape Olink to wide format
olink_wide <- olink %>%
  dplyr::select(SampleID, Assay, NPX) %>%
  pivot_wider(names_from = Assay, values_from = NPX)

# Merge with metadata
plot_data_all <- olink_wide %>%
  inner_join(metadata, by = c("SampleID" = "Barcode"))

cat("Plot data:", nrow(plot_data_all), "samples\n")

# Get recalculated FDR values
results_fdr <- results %>%
  filter(AveExpr > 0) %>%
  mutate(adj.P.Val = p.adjust(P.Value, method = "BH"))

# Function to plot one protein
plot_protein <- function(protein_name) {
  
  # Get FDR
  fdr_val <- results_fdr %>% filter(Assay == protein_name) %>% pull(adj.P.Val)
  fdr_label <- ifelse(length(fdr_val) > 0, paste0("FDR = ", signif(fdr_val, 2)), "")
  
  # Filter to R vs P only
  df <- plot_data_all %>%
    filter(Classification %in% c("Regressor", "Progressor"))
  
  ggplot(df, aes(x = Classification, y = .data[[protein_name]], fill = Classification)) +
    geom_violin(alpha = 0.3, color = NA, trim = FALSE) +
    geom_boxplot(width = 0.2, alpha = 0.9, outlier.shape = NA) +
    geom_jitter(width = 0.1, alpha = 0.6, size = 2) +
    scale_fill_manual(values = c("Progressor" = "#AF3B6E", "Regressor" = "#48639C")) +
    stat_compare_means(method = "wilcox.test", label.x.npc = 0.5, label.y.npc = 0.95) +
    labs(title = protein_name, subtitle = fdr_label, x = NULL, y = "NPX") +
    theme_classic() +
    theme(legend.position = "none",
          plot.title = element_text(hjust = 0.5, face = "bold"),
          plot.subtitle = element_text(hjust = 0.5, color = "gray40"))
}

# Get hit proteins
my_proteins <- hits_fdr$Assay

# Make all plots
plot_list <- lapply(my_proteins, plot_protein)
n_cols <- min(6, length(my_proteins))
n_rows <- ceiling(length(my_proteins) / n_cols)

combined <- plot_grid(plotlist = plot_list, ncol = n_cols)
ggsave(file.path(output_dir, "all_hits_boxplots.pdf"), combined)
print(combined)
```

---

# 6. KEGG Pathway Enrichment

```{r kegg-enrichment}
# =============================================================================
# PURPOSE: KEGG pathway enrichment
#
# KEGG: Curated pathways for metabolism, signaling, disease
# Often more interpretable than GO for mechanism-focused questions
# =============================================================================

kegg <- enrichKEGG(gene = entrez_ids$ENTREZID,
                   organism = "hsa",  # Human
                   pAdjustMethod = "BH",
                   pvalueCutoff = 0.05,
                   qvalueCutoff = 0.2)

cat("=== KEGG Pathways ===\n")
if (!is.null(kegg) && nrow(kegg) > 0) {
  cat("Significant pathways:", nrow(kegg), "\n")
  print(head(kegg@result[, c("Description", "GeneRatio", "pvalue", "p.adjust", "geneID")], 10))
} else {
  cat("No significant pathways found\n")
}
```

```{r kegg-plot, fig.width=10, fig.height=6}
if (!is.null(kegg) && nrow(kegg) > 0) {
  p2 <- dotplot(kegg, showCategory = 15, title = "KEGG Pathways")
  ggsave(file.path(output_dir, "KEGG_dotplot.pdf"), p2, width = 10, height = 6)
  print(p2)
}
```

---

# 7. Reactome Pathway Enrichment

```{r reactome-enrichment}
# =============================================================================
# PURPOSE: Reactome pathway enrichment
#
# Reactome: Peer-reviewed, highly curated pathway database
# Good for signaling cascades and detailed mechanistic pathways
# =============================================================================

reactome <- enrichPathway(gene = entrez_ids$ENTREZID,
                          organism = "human",
                          pAdjustMethod = "BH",
                          pvalueCutoff = 0.05,
                          qvalueCutoff = 0.2,
                          readable = TRUE)

cat("=== Reactome Pathways ===\n")
if (!is.null(reactome) && nrow(reactome) > 0) {
  cat("Significant pathways:", nrow(reactome), "\n")
  print(head(reactome@result[, c("Description", "GeneRatio", "pvalue", "p.adjust", "geneID")], 10))
} else {
  cat("No significant pathways found\n")
}
```

```{r reactome-plot, fig.width=10, fig.height=10}
if (!is.null(reactome) && nrow(reactome) > 0) {
  p3 <- dotplot(reactome, showCategory = 15, title = "Reactome Pathways")
  ggsave(file.path(output_dir, "Reactome_dotplot.pdf"), p3, width = 10, height = 10)
  print(p3)
}
```

---

# 8. MSigDB Hallmark Gene Sets

```{r hallmark}
# =============================================================================
# PURPOSE: Test enrichment against MSigDB Hallmark gene sets
#
# HALLMARK: 50 well-defined biological states/processes
# Examples: Inflammatory response, EMT, Hypoxia, TGF-beta signaling
# Very useful for high-level biological interpretation
# =============================================================================

# Get Hallmark gene sets
hallmark <- msigdbr(species = "Homo sapiens", category = "H") %>%
  dplyr::select(gs_name, entrez_gene)

# Run enrichment
hallmark_enrich <- enricher(gene = entrez_ids$ENTREZID,
                            TERM2GENE = hallmark,
                            pAdjustMethod = "BH",
                            pvalueCutoff = 0.05,
                            qvalueCutoff = 0.2)

cat("=== MSigDB Hallmark ===\n")
if (!is.null(hallmark_enrich) && nrow(hallmark_enrich) > 0) {
  cat("Significant gene sets:", nrow(hallmark_enrich), "\n")
  print(head(hallmark_enrich@result[, c("ID", "GeneRatio", "pvalue", "p.adjust")], 10))
} else {
  cat("No significant gene sets found\n")
}
```

```{r hallmark-plot, fig.width=10, fig.height=6}
if (!is.null(hallmark_enrich) && nrow(hallmark_enrich) > 0) {
  p4 <- dotplot(hallmark_enrich, showCategory = 15, title = "MSigDB Hallmark Gene Sets")
  ggsave(file.path(output_dir, "Hallmark_dotplot.pdf"), p4, width = 10, height = 6)
  print(p4)
}
```

---

# 9. Gene Set Enrichment Analysis (GSEA)

```{r gsea-prep}
# =============================================================================
# PURPOSE: GSEA uses ALL proteins ranked by fold change
#
# WHY GSEA vs ORA:
# - ORA: Uses only "significant" hits (information loss)
# - GSEA: Uses full ranked list (more sensitive, no threshold needed)
#
# RANKING: We use the signed statistic (t or logFC) so direction matters
# =============================================================================

# Create ranked gene list
# Merge results with Entrez IDs
results_entrez <- results %>%
  inner_join(all_entrez, by = c("Assay" = "SYMBOL"))

# Create named vector: names = Entrez IDs, values = ranking metric
# Using t-statistic (accounts for both effect size and variance)
gene_list <- results_entrez$t
names(gene_list) <- results_entrez$ENTREZID

# Sort by decreasing value (most positive first)
gene_list <- sort(gene_list, decreasing = TRUE)

cat("GSEA gene list:", length(gene_list), "genes\n")
cat("Range:", min(gene_list), "to", max(gene_list), "\n")
```

```{r gsea-go}
# GSEA with GO Biological Process
gsea_go <- gseGO(geneList = gene_list,
                 OrgDb = org.Hs.eg.db,
                 ont = "BP",
                 minGSSize = 10,
                 maxGSSize = 500,
                 pvalueCutoff = 0.05,
                 verbose = FALSE)

cat("=== GSEA: GO Biological Process ===\n")
if (!is.null(gsea_go) && nrow(gsea_go) > 0) {
  cat("Significant terms:", nrow(gsea_go), "\n")
  print(head(gsea_go@result[, c("Description", "NES", "pvalue", "p.adjust")], 10))
} else {
  cat("No significant terms found\n")
}
```

```{r gsea-go-plot, fig.width=10, fig.height=8}
if (!is.null(gsea_go) && nrow(gsea_go) > 0) {
  p_gsea_go <- ridgeplot(gsea_go, showCategory = 15) + 
    labs(title = "GSEA: GO Biological Process")
  ggsave(file.path(output_dir, "GSEA_GO_BP_ridgeplot.pdf"), p_gsea_go, width = 10, height = 8)
  print(p_gsea_go)
}
```

```{r gsea-hallmark}
# GSEA with Hallmark
gsea_hallmark <- GSEA(geneList = gene_list,
                      TERM2GENE = hallmark,
                      minGSSize = 10,
                      maxGSSize = 500,
                      pvalueCutoff = 0.05,
                      verbose = FALSE)

cat("\n=== GSEA: Hallmark ===\n")
if (!is.null(gsea_hallmark) && nrow(gsea_hallmark) > 0) {
  cat("Significant gene sets:", nrow(gsea_hallmark), "\n")
  print(head(gsea_hallmark@result[, c("ID", "NES", "pvalue", "p.adjust")], 10))
} else {
  cat("No significant gene sets found\n")
}
```

```{r gsea-hallmark-plot, fig.width=10, fig.height=6}
if (!is.null(gsea_hallmark) && nrow(gsea_hallmark) > 0) {
  p5 <- ridgeplot(gsea_hallmark, showCategory = 15) + 
    labs(title = "GSEA: Hallmark Gene Sets")
  ggsave(file.path(output_dir, "GSEA_Hallmark_ridgeplot.pdf"), p5, width = 10, height = 6)
  print(p5)
}
```

---

# 10. Direction-Specific Enrichment

```{r direction-ora, fig.height=10}
# =============================================================================
# PURPOSE: Separate ORA for up- and down-regulated proteins
# =============================================================================

# Split hits by direction
hits_up <- results %>%
  filter(AveExpr > 0) %>%
  mutate(adj.P.Val = p.adjust(P.Value, method = "BH")) %>%
  filter(adj.P.Val < 0.2) %>%
  filter(logFC > 0)  # Higher in Regressors

hits_down <- results %>%
  filter(AveExpr > 0) %>%
  mutate(adj.P.Val = p.adjust(P.Value, method = "BH")) %>%
  filter(adj.P.Val < 0.2) %>%
  filter(logFC < 0)  # Lower in Regressors (higher in Progressors)

cat("Higher in Regressors:", nrow(hits_up), "\n")
cat("Lower in Regressors:", nrow(hits_down), "\n")

# Convert to Entrez IDs
genes_up <- bitr(hits_up$Assay, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)
genes_down <- bitr(hits_down$Assay, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)

# Background (all detected proteins)
background <- results %>% filter(AveExpr > 0)
background_genes <- bitr(background$Assay, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)

# ORA: Higher in Regressors
if (nrow(genes_up) > 0) {
  ego_up <- enrichGO(
    gene = genes_up$ENTREZID,
    universe = background_genes$ENTREZID,
    OrgDb = org.Hs.eg.db,
    ont = "BP",
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05,
    readable = TRUE
  )
  
  cat("\n=== GO: Higher in Regressors ===\n")
  if (!is.null(ego_up) && nrow(ego_up) > 0) {
    print(head(ego_up@result[, c("Description", "GeneRatio", "pvalue", "p.adjust")], 10))
  }
}

# ORA: Lower in Regressors
if (nrow(genes_down) > 0) {
  ego_down <- enrichGO(
    gene = genes_down$ENTREZID,
    universe = background_genes$ENTREZID,
    OrgDb = org.Hs.eg.db,
    ont = "BP",
    pAdjustMethod = "BH",
    pvalueCutoff = 0.05,
    readable = TRUE
  )
  
  cat("\n=== GO: Lower in Regressors ===\n")
  if (!is.null(ego_down) && nrow(ego_down) > 0) {
    print(head(ego_down@result[, c("Description", "GeneRatio", "pvalue", "p.adjust")], 10))
  }
}
```

```{r direction-plots, fig.width=10, fig.height=10}
# Plots if results exist
if (exists("ego_up") && !is.null(ego_up) && nrow(ego_up) > 0) {
  p_up <- dotplot(ego_up, showCategory = 20, title = "GO BP: Higher in Regressors")
  ggsave(file.path(output_dir, "GO_BP_higher_in_Regressors.pdf"), p_up, width = 10, height = 10)
  print(p_up)
}

if (exists("ego_down") && !is.null(ego_down) && nrow(ego_down) > 0) {
  p_down <- dotplot(ego_down, showCategory = 15, title = "GO BP: Lower in Regressors (Higher in Progressors)")
  ggsave(file.path(output_dir, "GO_BP_lower_in_Regressors.pdf"), p_down, width = 10, height = 10)
  print(p_down)
}
```

---

# 11. Summary Table

```{r summary}
# =============================================================================
# PURPOSE: Combine all enrichment results into one summary
# =============================================================================

# Function to extract top results
extract_top <- function(result, source, n = 5) {
  if (is.null(result) || nrow(result) == 0) return(NULL)
  result@result %>%
    head(n) %>%
    mutate(Source = source) %>%
    dplyr::select(Source, ID, Description, pvalue, p.adjust)
}

# Combine results
summary_table <- bind_rows(
  extract_top(go_bp, "GO_BP"),
  extract_top(go_mf, "GO_MF"),
  extract_top(kegg, "KEGG"),
  extract_top(reactome, "Reactome"),
  extract_top(hallmark_enrich, "Hallmark")
)

print(summary_table)

# Save summary
write.csv(summary_table, file.path(output_dir, "enrichment_summary.csv"), row.names = FALSE)
```

---

# 12. Export Results

```{r export}
# Export all results
if (!is.null(go_bp) && nrow(go_bp) > 0) {
  write.csv(go_bp@result, file.path(output_dir, "enrichment_GO_BP.csv"), row.names = FALSE)
}
if (!is.null(go_mf) && nrow(go_mf) > 0) {
  write.csv(go_mf@result, file.path(output_dir, "enrichment_GO_MF.csv"), row.names = FALSE)
}
if (!is.null(go_cc) && nrow(go_cc) > 0) {
  write.csv(go_cc@result, file.path(output_dir, "enrichment_GO_CC.csv"), row.names = FALSE)
}
if (!is.null(kegg) && nrow(kegg) > 0) {
  write.csv(kegg@result, file.path(output_dir, "enrichment_KEGG.csv"), row.names = FALSE)
}
if (!is.null(reactome) && nrow(reactome) > 0) {
  write.csv(reactome@result, file.path(output_dir, "enrichment_Reactome.csv"), row.names = FALSE)
}
if (!is.null(hallmark_enrich) && nrow(hallmark_enrich) > 0) {
  write.csv(hallmark_enrich@result, file.path(output_dir, "enrichment_Hallmark.csv"), row.names = FALSE)
}
if (!is.null(gsea_go) && nrow(gsea_go) > 0) {
  write.csv(gsea_go@result, file.path(output_dir, "gsea_GO_BP.csv"), row.names = FALSE)
}
if (!is.null(gsea_hallmark) && nrow(gsea_hallmark) > 0) {
  write.csv(gsea_hallmark@result, file.path(output_dir, "gsea_Hallmark.csv"), row.names = FALSE)
}

# Export hit lists
write.csv(hits_fdr, file.path(output_dir, "protein_hits_FDR02.csv"), row.names = FALSE)
write.csv(hits_top20, file.path(output_dir, "protein_hits_top20.csv"), row.names = FALSE)

cat("Results exported to:", output_dir, "\n")
cat("\nFiles saved:\n")
list.files(output_dir)
```

---

# Session Info

```{r session-info}
sessionInfo()
```
