---
title: "SSc-ILD Onset: Identification of Deregulated Proteins"
author: "Analysis Script"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

This analysis identifies serum proteins associated with ILD onset in SSc patients.

**Analytical approach:**

| Analysis | Comparison | Description |
|----------|------------|-------------|
| Binary | ILD onset vs No ILD onset | Identifies proteins differentially expressed between groups |

**Covariates:**

- `entrytime` (year of study entry): Included to adjust for potential batch effects

---

# 1. Libraries

```{r libraries}
# =============================================================================
# PURPOSE: Load required R packages
#
# Data I/O:
#   - arrow: Read parquet files (Olink output format)
#   - readxl: Read Excel metadata
#   - here: Relative file paths for reproducibility
#
# Data manipulation:
#   - dplyr: Data wrangling (filter, mutate, select, etc.)
#   - tidyr: Reshaping data (long <-> wide format)
#   - tibble: For rownames_to_column conversion
#
# Statistical analysis:
#   - limma: Linear models for differential expression
#            Designed for microarray/proteomics; handles small samples well
#            Uses empirical Bayes to borrow information across proteins
#
# Visualization:
#   - ggplot2: General plotting
#   - EnhancedVolcano: Publication-quality volcano plots
#   - enrichR: Pathway enrichment analysis
# =============================================================================

library(arrow)
library(readxl)
library(here)
library(dplyr)
library(tidyr)
library(tibble)
library(limma)
library(ggplot2)
library(EnhancedVolcano)
library(enrichR)
```

---

# 2. Data Loading

```{r load-data}
# =============================================================================
# PURPOSE: Load raw data files
#
# Two data sources:
# 1. NPX data: Olink's Normalized Protein eXpression values (log2 scale)
#    - Already normalized across plates by Olink
#    - Long format: one row per protein-sample combination
#
# 2. Metadata: Clinical data including ILD onset status
# =============================================================================

# UPDATE THESE PATHS to your local files
olink <- read_parquet(here::here("data", "P2024-199-OLI_80_NPX_2024-10-10.parquet"))
metadata <- read_excel(here::here("data", "ASTRID_ILDonset_PROTEOMIC_COSIMO.xlsx"), sheet = 1)

cat("Olink data:", nrow(olink), "rows\n")
cat("Metadata:", nrow(metadata), "samples\n")
```

---

# 3. Data Preparation

```{r prepare-data}
# =============================================================================
# PURPOSE: Merge proteomics data with clinical metadata
#
# WHY: NPX data contains only sample IDs and protein values.
#      We need to link these to clinical outcomes (ILD onset)
#      to test which proteins associate with disease progression.
#
# KEY STEPS:
# 1. Create matching ID columns for merge
# 2. Remove unnecessary columns that cause issues
# 3. Merge NPX with metadata
# 4. Filter to samples with clinical data
# =============================================================================

# Remove columns that are not needed and may cause issues
metadata$`Id Visit V2018` <- NULL

# Create matching ID columns
# WHY: Olink uses "SampleID", metadata uses "Barcode" — same values, different names
metadata$patient <- metadata$Barcode
olink$patient <- olink$SampleID

# Merge NPX with metadata
merged <- merge(olink, metadata, by = "patient", all = TRUE)

# Keep only samples with clinical data (DPS = disease status) OR plate controls
# WHY: Plate controls (CTRL samples) are needed for QC but have no clinical data
#      Samples without DPS cannot be used in the analysis
filtered_df <- merged %>%
  filter(!is.na(DPS) | grepl("CTRL", SampleID))

cat("Merged data:", nrow(merged), "rows\n")
cat("Filtered data:", nrow(filtered_df), "rows\n")
```

```{r prepare-metadata}
# =============================================================================
# PURPOSE: Prepare clinical metadata for statistical modeling
#
# KEY STEPS:
# 1. Create ILD status variable from numeric indicator
# 2. Convert categorical variables to factors
# 3. Handle entry time (for batch effect adjustment)
# =============================================================================

# Create readable ILD status variable
# WHY: ILDonset_1yes0no is numeric (1/0); we need a factor for modeling
metadata$ild_status <- ifelse(metadata$ILDonset_1yes0no == 1, "ILDonset", "NoILDonset")

# Factor categorical variables for limma
# WHY: limma uses factor levels to define contrasts
metadata$ILDonset_1yes0no <- factor(metadata$ILDonset_1yes0no)
metadata$DPS <- factor(metadata$DPS)

# Extract entry year for batch effect adjustment
# WHY: Year of sample collection may introduce batch effects
#      (different reagent lots, operators, storage times)
metadata$entrytime <- substr(metadata$Eingangsdt., 1, 4)
metadata$entrytime <- factor(metadata$entrytime)

cat("ILD status distribution:\n")
print(table(metadata$ild_status))
cat("\nEntry year distribution:\n")
print(table(metadata$entrytime))
```

---

# 4. Expression Matrix Creation

```{r create-matrix}
# =============================================================================
# PURPOSE: Reshape NPX data from long to wide format for limma
#
# WHY: Olink outputs data in "long" format (one row per protein-sample)
#      limma requires a matrix: rows = proteins, columns = samples
#      This is the standard format for differential expression analysis
# =============================================================================

# Extract the three columns we need
matrix_df <- as.data.frame(cbind(
  filtered_df$patient, 
  filtered_df$NPX, 
  filtered_df$OlinkID
))
colnames(matrix_df) <- c("Patient", "Value", "Protein")

# Reshape: long → wide (proteins as rows, patients as columns)
wide_data <- tidyr::pivot_wider(matrix_df, names_from = Patient, values_from = Value)
wide_data <- as.data.frame(wide_data)

# Set protein IDs as rownames (limma convention)
rownames(wide_data) <- wide_data$Protein

# Remove non-sample columns (Protein column, plate controls, NA columns)
# WHY: We only want patient samples in the expression matrix
cols_to_remove <- c("Protein")
# Also identify plate control columns (contain "CTRL" or are NA)
ctrl_cols <- grep("CTRL|CONTROL", colnames(wide_data), value = TRUE)
na_cols <- colnames(wide_data)[colnames(wide_data) == "NA"]
cols_to_remove <- c(cols_to_remove, ctrl_cols, na_cols)

data <- wide_data[, !(colnames(wide_data) %in% cols_to_remove), drop = FALSE]

# Convert to numeric matrix
# WHY: pivot_wider creates character columns; limma needs numeric
data_matrix <- as.matrix(data)
data_matrix2 <- apply(data_matrix, 2, as.numeric)
rownames(data_matrix2) <- rownames(data_matrix)

cat("Expression matrix:", nrow(data_matrix2), "proteins x", ncol(data_matrix2), "samples\n")
```

---

# 5. Sample Alignment (CRITICAL)

```{r align-samples}
# =============================================================================
# PURPOSE: Ensure expression matrix columns match metadata rows
#
# WHY THIS IS CRITICAL:
# limma assumes row i of the design matrix corresponds to column i of the 
# expression matrix. If these are misaligned, you're testing the wrong 
# samples against the wrong clinical outcomes — results will be meaningless.
#
# This is the #1 source of errors in differential expression analysis!
# =============================================================================

# First, filter metadata to only include samples in the expression matrix
metadata_aligned <- metadata %>%
  filter(Barcode %in% colnames(data_matrix2))

# Check alignment BEFORE fixing
cat("BEFORE alignment:\n")
cat("Matrix columns (first 5):", head(colnames(data_matrix2), 5), "\n")
cat("Metadata Barcode (first 5):", head(metadata_aligned$Barcode, 5), "\n")
cat("Aligned:", identical(colnames(data_matrix2), metadata_aligned$Barcode), "\n\n")

# Reorder BOTH to ensure alignment
# Method: Sort both by the same ID
data_matrix2 <- data_matrix2[, order(colnames(data_matrix2))]
metadata_aligned <- metadata_aligned[order(metadata_aligned$Barcode), ]

# Verify alignment AFTER fixing
cat("AFTER alignment:\n")
cat("Matrix columns (first 5):", head(colnames(data_matrix2), 5), "\n")
cat("Metadata Barcode (first 5):", head(metadata_aligned$Barcode, 5), "\n")
cat("Aligned:", identical(colnames(data_matrix2), metadata_aligned$Barcode), "\n")

# STOP if not aligned — do not proceed with wrong results
stopifnot("Sample alignment failed!" = identical(colnames(data_matrix2), metadata_aligned$Barcode))
cat("\n✓ Alignment verified\n")
```

---
  
# 6. Binary Analysis: ILD Onset vs No ILD Onset

```{r binary-analysis}
# =============================================================================
# PURPOSE: Test each protein for differential expression between groups
#
# WHY LIMMA:
# - Designed for small sample sizes (borrows strength across proteins)
# - Empirical Bayes shrinkage stabilizes variance estimates
# - Well-validated for proteomics data
#
# MODEL: ~0 + ild_status + entrytime
# - "~0 + ild_status": Estimate mean for each group separately (no intercept)
#   This makes contrast specification cleaner
# - "+ entrytime": Adjust for year of entry (batch effect)
#
# CONTRAST: ILDonset - NoILDonset
# - Positive logFC = higher in ILD onset patients
# - Negative logFC = lower in ILD onset patients (higher in controls)
# =============================================================================

# Set factor with reference level
# WHY: "NoILDonset" first means positive logFC = higher in ILDonset
metadata_aligned$ild_status <- factor(metadata_aligned$ild_status, 
                                       levels = c("NoILDonset", "ILDonset"))

# Create design matrix
design <- model.matrix(~0 + ild_status + entrytime, data = metadata_aligned)

# Clean column names (remove "ild_status" prefix for readability)
colnames(design) <- gsub("ild_status", "", colnames(design))

cat("Design matrix:\n")
print(head(design))
cat("\nSamples per group:\n")
print(table(metadata_aligned$ild_status))

# Fit linear model
fit <- lmFit(data_matrix2, design)

# Define contrast: ILD onset vs No ILD onset
contrast_matrix <- makeContrasts(
  ILDonset_vs_NoILDonset = ILDonset - NoILDonset,
  levels = design
)

# Apply contrast and empirical Bayes moderation
# WHY eBayes: Shrinks variance estimates toward a common value
#             This stabilizes results when sample size is small
fit2 <- contrasts.fit(fit, contrast_matrix)
fit2 <- eBayes(fit2)

# Extract results
results <- topTable(fit2, coef = "ILDonset_vs_NoILDonset", number = Inf, adjust.method = "BH")

cat("\nTop 10 proteins:\n")
print(head(results, 10))
```

---

# 7. Expression Filtering & FDR Recalculation

```{r filter-results}
# =============================================================================
# PURPOSE: Filter for reliably detected proteins and recalculate FDR
#
# WHY FILTER ON AveExpr:
# - Proteins with very low expression (AveExpr < 0) are near detection limit
# - These have high noise and low statistical power
# - Removing them reduces multiple testing burden
#
# WHY RECALCULATE FDR:
# - Original adj.P.Val was calculated on ALL proteins
# - After filtering, we have fewer tests
# - Recalculating gives appropriate FDR for the filtered set
# - This is valid because AveExpr is INDEPENDENT of the test statistic
#   (it's mean expression across ALL samples, not related to group difference)
# =============================================================================

results <- results %>%
  rownames_to_column(var = "OlinkID")

# Filter for reliably detected proteins and recalculate FDR
results_filtered <- results %>%
  filter(AveExpr > 0) %>%
  mutate(adj.P.Val = p.adjust(P.Value, method = "BH"))

cat("Before filtering:", nrow(results), "proteins\n")
cat("After AveExpr > 0:", nrow(results_filtered), "proteins\n")

# Apply significance threshold
hits <- results_filtered %>%
  filter(adj.P.Val < 0.2) %>%
  arrange(P.Value)

cat("Significant hits (FDR < 0.2):", nrow(hits), "\n")
```

---

# 8. Add Gene Names

```{r add-gene-names}
# =============================================================================
# PURPOSE: Map Olink IDs to gene symbols for interpretability
#
# WHY: Olink IDs (e.g., "OID45450") are internal identifiers
#      Gene symbols (e.g., "MMP1") are more meaningful for biology
# =============================================================================

# Create lookup table from original data
gene_lookup <- filtered_df %>%
  distinct(OlinkID, Assay) %>%
  filter(!is.na(OlinkID))

# Add gene names to results
results_filtered <- results_filtered %>%
  left_join(gene_lookup, by = "OlinkID")

hits <- hits %>%
  left_join(gene_lookup, by = "OlinkID")

cat("Top hits with gene names:\n")
print(hits %>% dplyr::select(Assay, logFC, P.Value, adj.P.Val) %>% head(15))
```

---

# 9. Direction of Change Summary

```{r direction-summary}
# =============================================================================
# PURPOSE: Summarize which proteins are up/down in ILD onset
#
# INTERPRETATION:
# - Positive logFC: Higher in ILD onset patients
# - Negative logFC: Lower in ILD onset patients (higher in controls)
# =============================================================================

hits_up <- hits %>% filter(logFC > 0)
hits_down <- hits %>% filter(logFC < 0)

cat("=== Direction of Change (FDR < 0.2) ===\n\n")
cat("Higher in ILD onset:", nrow(hits_up), "\n")
if(nrow(hits_up) > 0) cat("  ", paste(hits_up$Assay, collapse = ", "), "\n\n")

cat("Lower in ILD onset:", nrow(hits_down), "\n")
if(nrow(hits_down) > 0) cat("  ", paste(hits_down$Assay, collapse = ", "), "\n")
```

---

# 10. Volcano Plot

```{r volcano-plot, fig.width=10, fig.height=10}
# =============================================================================
# PURPOSE: Visualize effect sizes and significance for all proteins
#
# X-axis: logFC (effect size)
#   - Positive = higher in ILD onset
#   - Negative = lower in ILD onset
#
# Y-axis: -log10(P.Value)
#   - Higher = more significant
#   - Note: Using nominal P.Value for visualization; FDR for hit selection
# =============================================================================

volcano <- EnhancedVolcano(
  results_filtered,
  lab = results_filtered$Assay,
  x = 'logFC',
  y = 'P.Value',
  pCutoff = 0.05,
  FCcutoff = 0.5,
  labSize = 3,
  drawConnectors = TRUE,
  widthConnectors = 0.5,
  title = "ILD Onset vs No ILD Onset",
  subtitle = "Positive logFC = higher in ILD onset patients"
)

print(volcano)
```

---

# 11. Boxplots for Significant Proteins

```{r boxplots, fig.width=12, fig.height=12}
# =============================================================================
# PURPOSE: Visualize expression of significant proteins
#
# WHY BOXPLOTS:
# - Show distribution in each group
# - Easy to interpret for non-statisticians
# - Wilcoxon p-value shown for binary comparison (different from limma)
#
# NOTE: Proteins were selected using limma (adjusts for entrytime);
#       Wilcoxon shown here is unadjusted binary comparison
# =============================================================================

# Prepare data for plotting (need long format with patient info)
plot_data <- filtered_df %>%
  filter(patient %in% metadata_aligned$Barcode) %>%
  left_join(metadata_aligned %>% dplyr::select(Barcode, ild_status), 
            by = c("patient" = "Barcode"))

# Function to plot one protein
plot_protein <- function(protein_id, gene_name, fdr_val) {
  
  df <- plot_data %>% filter(OlinkID == protein_id)
  
  ggplot(df, aes(x = ild_status, y = NPX, fill = ild_status)) +
    geom_boxplot(width = 0.5, alpha = 0.8, outlier.shape = NA) +
    geom_jitter(width = 0.15, alpha = 0.6, size = 2) +
    scale_fill_manual(values = c("NoILDonset" = "#48639C", "ILDonset" = "#AF3B6E")) +
    stat_compare_means(method = "wilcox.test", label.x.npc = 0.5, label.y.npc = 0.95) +
    labs(
      title = gene_name,
      subtitle = paste0("FDR = ", signif(fdr_val, 2)),
      x = NULL,
      y = "NPX"
    ) +
    theme_classic() +
    theme(
      legend.position = "none",
      plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
      plot.subtitle = element_text(hjust = 0.5, size = 10, color = "gray40")
    )
}

# Plot top hits (up to 12)
if(nrow(hits) > 0) {
  library(cowplot)
  library(ggpubr)
  
  n_plots <- min(nrow(hits), 12)
  plot_list <- list()
  
  for(i in 1:n_plots) {
    plot_list[[i]] <- plot_protein(
      hits$OlinkID[i], 
      hits$Assay[i], 
      hits$adj.P.Val[i]
    )
  }
  
  combined <- plot_grid(plotlist = plot_list, ncol = 4)
  print(combined)
}
```

---

# 12. Pathway Enrichment

```{r enrichment, fig.width=10, fig.height=8}
# =============================================================================
# PURPOSE: Identify biological pathways enriched in significant proteins
#
# METHOD: Over-representation analysis (ORA) using Enrichr
#
# DATABASES:
# - GO Biological Process: Cellular processes
# - GO Molecular Function: Biochemical activities
# - GO Cellular Component: Subcellular localization
#
# NOTE: With small protein lists, enrichment may not reach significance
#       This is normal for targeted panels (vs whole proteome)
# =============================================================================

if(nrow(hits) >= 3) {
  
  proteins <- hits$Assay
  
  dbs <- c("GO_Biological_Process_2023", 
           "GO_Molecular_Function_2023", 
           "GO_Cellular_Component_2023")
  
  enriched <- enrichr(proteins, dbs)
  
  # Show top GO Biological Process results
  bp_results <- enriched[["GO_Biological_Process_2023"]]
  
  if(nrow(bp_results) > 0 && any(bp_results$Adjusted.P.value < 0.1)) {
    plotEnrich(bp_results, 
               showTerms = 15, 
               numChar = 60, 
               y = "Count", 
               orderBy = "Adjusted.P.value", 
               title = "GO Biological Process Enrichment")
  } else {
    cat("No significant GO enrichment found (FDR < 0.1)\n")
    cat("This is common with small, targeted protein panels.\n")
    cat("\nTop terms (may not be significant):\n")
    print(head(bp_results[, c("Term", "Overlap", "Adjusted.P.value")], 10))
  }
  
} else {
  cat("Too few hits for enrichment analysis (need at least 3 proteins)\n")
}
```

---

# 13. Export Results

```{r export}
# =============================================================================
# PURPOSE: Save results for downstream analysis and reporting
# =============================================================================

output_dir <- here::here("output", "ILD_Onset_Analysis")
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

# Full results (all proteins)
write.csv(results_filtered, 
          file.path(output_dir, "results_ILDonset_vs_NoILDonset_all.csv"), 
          row.names = FALSE)

# Significant hits only
write.csv(hits, 
          file.path(output_dir, "results_ILDonset_vs_NoILDonset_FDR02.csv"), 
          row.names = FALSE)

cat("Results exported to:", output_dir, "\n")
cat("Files:\n")
cat("  - results_ILDonset_vs_NoILDonset_all.csv (all proteins)\n")
cat("  - results_ILDonset_vs_NoILDonset_FDR02.csv (FDR < 0.2 hits)\n")
```

---

# Session Info

```{r session-info}
sessionInfo()
```
