---
title: "SSc Proteomics: Pathway Enrichment Analysis"
author: "Analysis Script"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

Pathway enrichment analysis on protein hits from the **ILD Onset vs No ILD Onset** binary analysis.

**Hit selection:**
- Filter for AveExpr > 0 (reliably detected proteins)
- Recalculate FDR on filtered set
- Apply FDR < 0.2 threshold

**Enrichment approaches:**
1. **All hits** — combined analysis
2. **Up in ILD onset** — proteins higher in patients who develop ILD
3. **Down in ILD onset** — proteins lower in patients who develop ILD

```{r libraries}
# =============================================================================
# REQUIRED PACKAGES
# Install if needed:
#   BiocManager::install(c("clusterProfiler", "org.Hs.eg.db", "enrichplot", "ReactomePA"))
#   install.packages("msigdbr")
# =============================================================================

library(dplyr)
library(ggplot2)
library(tidyr)

# Enrichment analysis
library(clusterProfiler)    # GO, KEGG enrichment
library(org.Hs.eg.db)       # Human gene annotation
library(enrichplot)         # Visualization
library(ReactomePA)         # Reactome pathway analysis

# Gene sets
library(msigdbr)            # MSigDB gene sets (Hallmark, etc.)
```

---

# 1. Load Results

```{r load-results}
# =============================================================================
# PURPOSE: Load results from the ILD onset binary analysis
#
# This should be the output from your main analysis script
# (results_ILDonset_vs_NoILDonset_all.csv)
# =============================================================================

# Load the binary comparison results
results <- read.csv(here::here("output", "ILD_Onset_Analysis", "results_ILDonset_vs_NoILDonset_all.csv"))

cat("Loaded", nrow(results), "proteins\n")
head(results)
```

---

# 2. Define Protein Hits

```{r define-hits}
# =============================================================================
# PURPOSE: Define significant proteins
#
# STRATEGY:
# 1. Filter for reliably detected proteins (AveExpr > 0)
# 2. Recalculate FDR on filtered set (valid because AveExpr is independent)
# 3. Apply FDR < 0.2 threshold
#
# WHY RECALCULATE FDR:
# - Original adj.P.Val was calculated on ALL proteins
# - After filtering, fewer tests → less harsh correction → better power
# =============================================================================

# Filter and recalculate FDR
hits_fdr <- results %>%
  filter(AveExpr > 0) %>%
  mutate(adj.P.Val = p.adjust(P.Value, method = "BH")) %>%
  filter(adj.P.Val < 0.2) %>%
  arrange(P.Value)

cat("=== FDR < 0.2 (after AveExpr > 0 filter) ===\n")
cat("N proteins:", nrow(hits_fdr), "\n")
if(nrow(hits_fdr) > 0) {
  cat("Proteins:", paste(hits_fdr$Assay, collapse = ", "), "\n")
}
```

```{r direction-of-change}
# =============================================================================
# PURPOSE: Separate up- and down-regulated proteins
#
# WHY: Proteins higher vs lower in ILD onset may reflect different biology
#
# INTERPRETATION (from binary analysis):
# - Positive logFC: Higher in ILD onset patients
# - Negative logFC: Lower in ILD onset patients (higher in controls)
# =============================================================================

hits_up <- hits_fdr %>% filter(logFC > 0)
hits_down <- hits_fdr %>% filter(logFC < 0)

cat("=== Direction of Change (FDR < 0.2) ===\n\n")
cat("Higher in ILD onset (positive logFC):", nrow(hits_up), "\n")
if(nrow(hits_up) > 0) cat("  ", paste(hits_up$Assay, collapse = ", "), "\n\n")

cat("Lower in ILD onset (negative logFC):", nrow(hits_down), "\n")
if(nrow(hits_down) > 0) cat("  ", paste(hits_down$Assay, collapse = ", "), "\n")
```

---

# 3. Gene ID Conversion

```{r convert-ids}
# =============================================================================
# PURPOSE: Convert gene symbols to Entrez IDs
#
# WHY: clusterProfiler and most pathway databases use Entrez IDs
#      Gene symbols (like "MMP1") must be converted
#
# NOTE: Some genes may not map — this is normal (pseudogenes, novel genes)
# =============================================================================

if(nrow(hits_fdr) >= 3) {
  
  gene_symbols <- hits_fdr$Assay
  
  entrez_ids <- bitr(gene_symbols, 
                     fromType = "SYMBOL", 
                     toType = "ENTREZID", 
                     OrgDb = org.Hs.eg.db)
  
  cat("Mapped", nrow(entrez_ids), "of", length(gene_symbols), "genes to Entrez IDs\n")
  
  # Check which genes failed to map
  unmapped <- setdiff(gene_symbols, entrez_ids$SYMBOL)
  if (length(unmapped) > 0) {
    cat("Unmapped genes:", paste(unmapped, collapse = ", "), "\n")
  }
  
  # Also map direction-specific lists
  if(nrow(hits_up) >= 2) {
    entrez_up <- bitr(hits_up$Assay, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)
    cat("\nMapped UP genes:", nrow(entrez_up), "\n")
  }
  
  if(nrow(hits_down) >= 2) {
    entrez_down <- bitr(hits_down$Assay, fromType = "SYMBOL", toType = "ENTREZID", OrgDb = org.Hs.eg.db)
    cat("Mapped DOWN genes:", nrow(entrez_down), "\n")
  }
  
} else {
  cat("Too few hits for gene ID conversion (need at least 3)\n")
}

# Create mapping for all results (for GSEA later)
results_detected <- results %>% filter(AveExpr > 0)
all_entrez <- bitr(results_detected$Assay, 
                   fromType = "SYMBOL", 
                   toType = "ENTREZID", 
                   OrgDb = org.Hs.eg.db)

cat("\nTotal mapped for GSEA:", nrow(all_entrez), "of", nrow(results_detected), "\n")
```

---

# 4. GO Enrichment Analysis (All Hits)

```{r go-enrichment}
# =============================================================================
# PURPOSE: Gene Ontology enrichment analysis
#
# THREE ONTOLOGIES:
# - BP (Biological Process): What the proteins do together
# - MF (Molecular Function): Biochemical activities
# - CC (Cellular Component): Where proteins localize
#
# METHOD: Over-representation analysis (ORA)
# Tests if hit genes are enriched in GO terms vs background
# =============================================================================

if(exists("entrez_ids") && nrow(entrez_ids) >= 3) {
  
  # GO Biological Process
  go_bp <- enrichGO(gene = entrez_ids$ENTREZID,
                    OrgDb = org.Hs.eg.db,
                    ont = "BP",
                    pAdjustMethod = "BH",
                    pvalueCutoff = 0.05,
                    qvalueCutoff = 0.2,
                    readable = TRUE)
  
  cat("=== GO Biological Process ===\n")
  if (!is.null(go_bp) && nrow(go_bp) > 0) {
    cat("Significant terms:", nrow(go_bp), "\n")
    print(head(go_bp@result[, c("Description", "GeneRatio", "pvalue", "p.adjust", "geneID")], 10))
  } else {
    cat("No significant terms found\n")
  }
  
} else {
  cat("Too few mapped genes for GO enrichment\n")
  go_bp <- NULL
}
```

```{r go-mf-cc}
if(exists("entrez_ids") && nrow(entrez_ids) >= 3) {
  
  # GO Molecular Function
  go_mf <- enrichGO(gene = entrez_ids$ENTREZID,
                    OrgDb = org.Hs.eg.db,
                    ont = "MF",
                    pAdjustMethod = "BH",
                    pvalueCutoff = 0.05,
                    qvalueCutoff = 0.2,
                    readable = TRUE)
  
  cat("\n=== GO Molecular Function ===\n")
  if (!is.null(go_mf) && nrow(go_mf) > 0) {
    cat("Significant terms:", nrow(go_mf), "\n")
    print(head(go_mf@result[, c("Description", "GeneRatio", "pvalue", "p.adjust")], 10))
  } else {
    cat("No significant terms found\n")
  }
  
  # GO Cellular Component
  go_cc <- enrichGO(gene = entrez_ids$ENTREZID,
                    OrgDb = org.Hs.eg.db,
                    ont = "CC",
                    pAdjustMethod = "BH",
                    pvalueCutoff = 0.05,
                    qvalueCutoff = 0.2,
                    readable = TRUE)
  
  cat("\n=== GO Cellular Component ===\n")
  if (!is.null(go_cc) && nrow(go_cc) > 0) {
    cat("Significant terms:", nrow(go_cc), "\n")
    print(head(go_cc@result[, c("Description", "GeneRatio", "pvalue", "p.adjust")], 10))
  } else {
    cat("No significant terms found\n")
  }
  
} else {
  go_mf <- NULL
  go_cc <- NULL
}
```

```{r go-plots, fig.width=10, fig.height=8}
# =============================================================================
# PURPOSE: Visualize GO enrichment results
# =============================================================================

if (!is.null(go_bp) && nrow(go_bp) > 0) {
  p1 <- dotplot(go_bp, showCategory = 15, title = "GO Biological Process")
  print(p1)
}
```

---

# 5. Direction-Specific Enrichment

```{r direction-enrichment}
# =============================================================================
# PURPOSE: Separate enrichment for UP and DOWN proteins
#
# WHY: Proteins elevated in ILD onset may reflect different pathways
#      than proteins depleted in ILD onset
# =============================================================================

# UP in ILD onset
if(exists("entrez_up") && nrow(entrez_up) >= 3) {
  
  go_bp_up <- enrichGO(gene = entrez_up$ENTREZID,
                       OrgDb = org.Hs.eg.db,
                       ont = "BP",
                       pAdjustMethod = "BH",
                       pvalueCutoff = 0.05,
                       qvalueCutoff = 0.2,
                       readable = TRUE)
  
  cat("=== GO BP: Higher in ILD Onset ===\n")
  if (!is.null(go_bp_up) && nrow(go_bp_up) > 0) {
    cat("Significant terms:", nrow(go_bp_up), "\n")
    print(head(go_bp_up@result[, c("Description", "GeneRatio", "pvalue", "p.adjust")], 10))
  } else {
    cat("No significant terms found\n")
  }
  
} else {
  cat("Too few UP genes for enrichment\n")
  go_bp_up <- NULL
}

# DOWN in ILD onset
if(exists("entrez_down") && nrow(entrez_down) >= 3) {
  
  go_bp_down <- enrichGO(gene = entrez_down$ENTREZID,
                         OrgDb = org.Hs.eg.db,
                         ont = "BP",
                         pAdjustMethod = "BH",
                         pvalueCutoff = 0.05,
                         qvalueCutoff = 0.2,
                         readable = TRUE)
  
  cat("\n=== GO BP: Lower in ILD Onset ===\n")
  if (!is.null(go_bp_down) && nrow(go_bp_down) > 0) {
    cat("Significant terms:", nrow(go_bp_down), "\n")
    print(head(go_bp_down@result[, c("Description", "GeneRatio", "pvalue", "p.adjust")], 10))
  } else {
    cat("No significant terms found\n")
  }
  
} else {
  cat("Too few DOWN genes for enrichment\n")
  go_bp_down <- NULL
}
```

```{r direction-plots, fig.width=12, fig.height=6}
if (!is.null(go_bp_up) && nrow(go_bp_up) > 0) {
  p_up <- dotplot(go_bp_up, showCategory = 15, title = "GO BP: Higher in ILD Onset")
  print(p_up)
}

if (!is.null(go_bp_down) && nrow(go_bp_down) > 0) {
  p_down <- dotplot(go_bp_down, showCategory = 15, title = "GO BP: Lower in ILD Onset")
  print(p_down)
}
```

---

# 6. KEGG Pathway Enrichment

```{r kegg-enrichment}
# =============================================================================
# PURPOSE: KEGG pathway enrichment
#
# KEGG: Curated pathways for metabolism, signaling, disease
# Often more interpretable than GO for mechanism-focused questions
# =============================================================================

if(exists("entrez_ids") && nrow(entrez_ids) >= 3) {
  
  kegg <- enrichKEGG(gene = entrez_ids$ENTREZID,
                     organism = "hsa",
                     pAdjustMethod = "BH",
                     pvalueCutoff = 0.05,
                     qvalueCutoff = 0.2)
  
  cat("=== KEGG Pathways ===\n")
  if (!is.null(kegg) && nrow(kegg) > 0) {
    cat("Significant pathways:", nrow(kegg), "\n")
    print(head(kegg@result[, c("Description", "GeneRatio", "pvalue", "p.adjust", "geneID")], 10))
  } else {
    cat("No significant pathways found\n")
  }
  
} else {
  kegg <- NULL
}
```

```{r kegg-plot, fig.width=10, fig.height=6}
if (!is.null(kegg) && nrow(kegg) > 0) {
  p2 <- dotplot(kegg, showCategory = 15, title = "KEGG Pathways")
  print(p2)
}
```

---

# 7. Reactome Pathway Enrichment

```{r reactome-enrichment}
# =============================================================================
# PURPOSE: Reactome pathway enrichment
#
# Reactome: Peer-reviewed, highly curated pathway database
# Good for signaling cascades and detailed mechanistic pathways
# =============================================================================

if(exists("entrez_ids") && nrow(entrez_ids) >= 3) {
  
  reactome <- enrichPathway(gene = entrez_ids$ENTREZID,
                            organism = "human",
                            pAdjustMethod = "BH",
                            pvalueCutoff = 0.05,
                            qvalueCutoff = 0.2,
                            readable = TRUE)
  
  cat("=== Reactome Pathways ===\n")
  if (!is.null(reactome) && nrow(reactome) > 0) {
    cat("Significant pathways:", nrow(reactome), "\n")
    print(head(reactome@result[, c("Description", "GeneRatio", "pvalue", "p.adjust", "geneID")], 10))
  } else {
    cat("No significant pathways found\n")
  }
  
} else {
  reactome <- NULL
}
```

```{r reactome-plot, fig.width=10, fig.height=6}
if (!is.null(reactome) && nrow(reactome) > 0) {
  p3 <- dotplot(reactome, showCategory = 15, title = "Reactome Pathways")
  print(p3)
}
```

---

# 8. MSigDB Hallmark Gene Sets

```{r hallmark}
# =============================================================================
# PURPOSE: Test enrichment against MSigDB Hallmark gene sets
#
# HALLMARK: 50 well-defined biological states/processes
# Examples: Inflammatory response, EMT, Hypoxia, TGF-beta signaling
# Very useful for high-level biological interpretation
# =============================================================================

if(exists("entrez_ids") && nrow(entrez_ids) >= 3) {
  
  # Get Hallmark gene sets
  hallmark <- msigdbr(species = "Homo sapiens", category = "H") %>%
    dplyr::select(gs_name, entrez_gene)
  
  # Run enrichment
  hallmark_enrich <- enricher(gene = entrez_ids$ENTREZID,
                              TERM2GENE = hallmark,
                              pAdjustMethod = "BH",
                              pvalueCutoff = 0.05,
                              qvalueCutoff = 0.2)
  
  cat("=== MSigDB Hallmark ===\n")
  if (!is.null(hallmark_enrich) && nrow(hallmark_enrich) > 0) {
    cat("Significant gene sets:", nrow(hallmark_enrich), "\n")
    print(head(hallmark_enrich@result[, c("ID", "GeneRatio", "pvalue", "p.adjust")], 10))
  } else {
    cat("No significant gene sets found\n")
  }
  
} else {
  hallmark_enrich <- NULL
}
```

```{r hallmark-plot, fig.width=10, fig.height=6}
if (!is.null(hallmark_enrich) && nrow(hallmark_enrich) > 0) {
  p4 <- dotplot(hallmark_enrich, showCategory = 15, title = "MSigDB Hallmark Gene Sets")
  print(p4)
}
```

---

# 9. Gene Set Enrichment Analysis (GSEA)

```{r gsea-prep}
# =============================================================================
# PURPOSE: GSEA uses ALL proteins ranked by fold change
#
# WHY GSEA vs ORA:
# - ORA: Uses only "significant" hits (information loss)
# - GSEA: Uses full ranked list (more sensitive, no threshold needed)
#
# RANKING: We use the t-statistic (accounts for effect size and variance)
# - Positive t = higher in ILD onset
# - Negative t = lower in ILD onset
# =============================================================================

# Create ranked gene list
# Merge results with Entrez IDs
results_entrez <- results_detected %>%
  inner_join(all_entrez, by = c("Assay" = "SYMBOL"))

# Create named vector: names = Entrez IDs, values = ranking metric
# Using t-statistic
gene_list <- results_entrez$t
names(gene_list) <- results_entrez$ENTREZID

# Sort by decreasing value (most positive = most elevated in ILD onset)
gene_list <- sort(gene_list, decreasing = TRUE)

cat("GSEA gene list:", length(gene_list), "genes\n")
cat("Range:", round(min(gene_list), 2), "to", round(max(gene_list), 2), "\n")
```

```{r gsea-go}
# GSEA with GO Biological Process
gsea_go <- gseGO(geneList = gene_list,
                 OrgDb = org.Hs.eg.db,
                 ont = "BP",
                 minGSSize = 10,
                 maxGSSize = 500,
                 pvalueCutoff = 0.05,
                 verbose = FALSE)

cat("=== GSEA: GO Biological Process ===\n")
if (!is.null(gsea_go) && nrow(gsea_go) > 0) {
  cat("Significant terms:", nrow(gsea_go), "\n")
  # Positive NES = enriched in ILD onset; Negative NES = depleted in ILD onset
  print(head(gsea_go@result[, c("Description", "NES", "pvalue", "p.adjust")], 10))
} else {
  cat("No significant terms found\n")
}
```

```{r gsea-hallmark}
# Get Hallmark gene sets if not already loaded
if(!exists("hallmark")) {
  hallmark <- msigdbr(species = "Homo sapiens", category = "H") %>%
    dplyr::select(gs_name, entrez_gene)
}

# GSEA with Hallmark
gsea_hallmark <- GSEA(geneList = gene_list,
                      TERM2GENE = hallmark,
                      minGSSize = 10,
                      maxGSSize = 500,
                      pvalueCutoff = 0.05,
                      verbose = FALSE)

cat("\n=== GSEA: Hallmark ===\n")
if (!is.null(gsea_hallmark) && nrow(gsea_hallmark) > 0) {
  cat("Significant gene sets:", nrow(gsea_hallmark), "\n")
  # Positive NES = enriched in ILD onset; Negative NES = depleted
  print(head(gsea_hallmark@result[, c("ID", "NES", "pvalue", "p.adjust")], 10))
} else {
  cat("No significant gene sets found\n")
}
```

```{r gsea-plot, fig.width=10, fig.height=6}
if (!is.null(gsea_hallmark) && nrow(gsea_hallmark) > 0) {
  # Ridge plot shows distribution of genes in each pathway
  p5 <- ridgeplot(gsea_hallmark, showCategory = 15) + 
    labs(title = "GSEA: Hallmark Gene Sets",
         subtitle = "Positive NES = enriched in ILD onset")
  print(p5)
}
```

---

# 10. Summary Table

```{r summary}
# =============================================================================
# PURPOSE: Combine all enrichment results into one summary
# =============================================================================

# Function to extract top results
extract_top <- function(result, source, n = 5) {
  if (is.null(result) || nrow(result) == 0) return(NULL)
  result@result %>%
    head(n) %>%
    mutate(Source = source) %>%
    dplyr::select(Source, ID, Description, pvalue, p.adjust)
}

# Combine results
summary_table <- bind_rows(
  extract_top(go_bp, "GO_BP"),
  extract_top(go_mf, "GO_MF"),
  extract_top(kegg, "KEGG"),
  extract_top(reactome, "Reactome"),
  extract_top(hallmark_enrich, "Hallmark")
)

summary_table
```

---

# 11. Export Results

```{r export}
# =============================================================================
# PURPOSE: Save results for downstream analysis and reporting
# =============================================================================

output_dir <- "enrichment_results_ILD"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

# ORA results
if (!is.null(go_bp) && nrow(go_bp) > 0) {
  write.csv(go_bp@result, file.path(output_dir, "ORA_GO_BP.csv"), row.names = FALSE)
}
if (!is.null(go_mf) && nrow(go_mf) > 0) {
  write.csv(go_mf@result, file.path(output_dir, "ORA_GO_MF.csv"), row.names = FALSE)
}
if (!is.null(go_cc) && nrow(go_cc) > 0) {
  write.csv(go_cc@result, file.path(output_dir, "ORA_GO_CC.csv"), row.names = FALSE)
}
if (!is.null(kegg) && nrow(kegg) > 0) {
  write.csv(kegg@result, file.path(output_dir, "ORA_KEGG.csv"), row.names = FALSE)
}
if (!is.null(reactome) && nrow(reactome) > 0) {
  write.csv(reactome@result, file.path(output_dir, "ORA_Reactome.csv"), row.names = FALSE)
}
if (!is.null(hallmark_enrich) && nrow(hallmark_enrich) > 0) {
  write.csv(hallmark_enrich@result, file.path(output_dir, "ORA_Hallmark.csv"), row.names = FALSE)
}

# Direction-specific ORA
if (exists("go_bp_up") && !is.null(go_bp_up) && nrow(go_bp_up) > 0) {
  write.csv(go_bp_up@result, file.path(output_dir, "ORA_GO_BP_UpInILD.csv"), row.names = FALSE)
}
if (exists("go_bp_down") && !is.null(go_bp_down) && nrow(go_bp_down) > 0) {
  write.csv(go_bp_down@result, file.path(output_dir, "ORA_GO_BP_DownInILD.csv"), row.names = FALSE)
}

# GSEA results
if (!is.null(gsea_go) && nrow(gsea_go) > 0) {
  write.csv(gsea_go@result, file.path(output_dir, "GSEA_GO_BP.csv"), row.names = FALSE)
}
if (!is.null(gsea_hallmark) && nrow(gsea_hallmark) > 0) {
  write.csv(gsea_hallmark@result, file.path(output_dir, "GSEA_Hallmark.csv"), row.names = FALSE)
}

# Hit lists
write.csv(hits_fdr, file.path(output_dir, "protein_hits_FDR02.csv"), row.names = FALSE)
write.csv(hits_up, file.path(output_dir, "protein_hits_UpInILD.csv"), row.names = FALSE)
write.csv(hits_down, file.path(output_dir, "protein_hits_DownInILD.csv"), row.names = FALSE)

cat("Results exported to:", output_dir, "\n")
```

---

# Session Info

```{r session-info}
sessionInfo()
```
