---
title: "Analysis Types: Binary vs. Continuous"
author: "Analysis Script"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, dpi = 300)

# =============================================================================
# OUTPUT DIRECTORY: output/<script_name>/
# =============================================================================

script_name <- "SSc_Skin_Regression_Analysis"

output_dir <- here::here("output", script_name)
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

cat("Output directory:", output_dir, "\n")

# Color palettes
colours_eight <- c("#a4161a", "#F3722C", "#F9C74F", "#90BE6D", "#43AA8B", "#4D908E", "#3d348b", "#277DA1")
colors_class <- c("Regressor" = "#48639C", "Non_Regressor" = "#AF3B6E", "Stable" = "#ECA72C", "Healthy" = "#84A07C")

```

## Overview

This analysis identifies serum proteins associated with skin fibrosis regression in SSc patients.

**Four analytical approaches:**

| Analysis | Comparison | Description |
|----------|------------|-------------|
| A | Binary: Regressor vs Non_Regressor | Includes Stable patients with Non_Regressors |
| B | Continuous: All patients | Correlation with ΔMRSS (degree of skin change) |
| C | Binary: Regressor vs Progressor | Excludes Stable — compares extremes only |
| D | Continuous: Regressor vs Progressor | Correlation with ΔMRSS, excludes Stable |

**Sensitivity analysis** restricts to years with both groups represented (2014, 2019-2021).

**Healthy controls** are included for visualization/reference but not in statistical models.

```{r libraries}
# Data I/O
library(arrow)        # Read parquet files (Olink output format)
library(openxlsx)     # Read Excel metadata

# Data manipulation
library(dplyr)        # Data wrangling
library(tidyr)        # Reshaping data (long <-> wide)

# Statistical analysis
library(limma)        # Linear models for differential expression
                      # Designed for microarray/proteomics; handles small samples well
                      # Uses empirical Bayes to borrow information across proteins

# Visualization
library(ggplot2)      # Plotting
library(ggpubr)       # Add statistical comparisons to plots
library(cowplot)      # Combine multiple plots
library(ggrepel)      # Non-overlapping labels on volcano plots
```

---

# 1. Data Loading

```{r load-data}
# =============================================================================
# PURPOSE: Load raw data files
#
# Two data sources:
# 1. NPX data: Olink's Normalized Protein eXpression values (log2 scale)
#    - Already normalized across plates by Olink
#    - Long format: one row per protein-sample combination
#
# 2. Metadata: Clinical data including mRSS scores and patient classification
# =============================================================================

# UPDATE THESE PATHS to your local files
npx_data <- read_parquet(here::here("data", "P2024-199-OLI_80_NPX_2024-10-10.parquet"))
metadata <- read.xlsx(here::here("data", "metadata.xlsx"))

cat("NPX data:", nrow(npx_data), "rows\n")
cat("Metadata:", nrow(metadata), "samples\n")
```

---

# 2. Data Preparation

```{r prepare-data}
# =============================================================================
# PURPOSE: Merge proteomics data with clinical metadata
#
# WHY: NPX data contains only sample IDs and protein values.
#      We need to link these to clinical outcomes (mRSS change, classification)
#      to test which proteins associate with skin regression.
# =============================================================================

# Create matching ID columns for merge
# (NPX uses "SampleID", metadata uses "Barcode" — same values, different names)
metadata$patient <- metadata$Barcode
npx_data$patient <- npx_data$SampleID

# Create binary grouping variable: Regressor vs Non_Regressor
# WHY: Stable patients didn't improve but also didn't worsen.
#      For the primary analysis, we group them with Progressors as "Non_Regressor"
#      to answer: "What distinguishes patients who improved from those who didn't?"
metadata$PatientGroup <- ifelse(
  metadata$Classification %in% c("Progressor", "Stable"),
  "Non_Regressor",
  metadata$Classification
)

# Merge NPX with metadata (left join: keep all NPX rows)
merged <- merge(npx_data, metadata, by = "patient", all.x = TRUE)
merged <- tibble::as_tibble(merged)

# Keep only samples with clinical outcome data OR plate controls
# WHY: Plate controls (CTRL samples) are needed for QC but have no clinical data
#      Samples without MRSS_Delta cannot be used in the analysis
filtered_df <- merged %>%
  filter(!is.na(MRSS_Delta) | grepl("CTRL", SampleID))

cat("Merged data:", nrow(merged), "rows\n")
cat("Filtered data:", nrow(filtered_df), "rows\n")
```

```{r prepare-metadata}
# =============================================================================
# PURPOSE: Prepare clinical metadata for statistical modeling
#
# KEY STEPS:
# 1. Exclude patients missing from the expression matrix
# 2. Calculate the continuous outcome variable (signed MRSS change)
# 3. Convert categorical variables to factors with proper reference levels
# =============================================================================

# Exclude patients 180 and 216
# WHY: These patients are in metadata but missing from NPX data (no proteomics)
filtered_metadata <- metadata %>%
  filter(!EUSTAR_Number %in% c(180, 216)) %>%
  filter(!is.na(MRSS_Delta))  # Must have clinical outcome

# Calculate SIGNED mRSS change: Baseline minus Follow-up
# WHY: MRSS_Delta in metadata is absolute value (unsigned)
#      We need direction: positive = improvement, negative = worsening
#      This enables the continuous correlation analysis
filtered_metadata$mRSS_FUP <- as.numeric(filtered_metadata$mRSS_FUP)
filtered_metadata$MRSS_signed <- filtered_metadata$mRSS_BL - filtered_metadata$mRSS_FUP

# Convert Visit_Year to factor
# WHY: Year of sample collection may introduce batch effects
#      (different reagent lots, operators, storage times)
#      We'll include it as a covariate to adjust for these effects
filtered_metadata$Visit_Year <- as.factor(filtered_metadata$Visit_Year)

# Set reference level for PatientGroup
# WHY: In limma, the first factor level is the reference (denominator in contrast)
#      Setting "Non_Regressor" first means positive logFC = higher in Regressor
filtered_metadata$PatientGroup <- factor(filtered_metadata$PatientGroup,
                                          levels = c("Non_Regressor", "Regressor"))

# Extract patient NPX data only (exclude plate controls)
# WHY: Controls are useful for QC but shouldn't be in statistical models
patient_npx <- filtered_df %>%
  filter(!grepl("CTRL", SampleID)) %>%
  filter(EUSTAR_Number %in% filtered_metadata$EUSTAR_Number)

cat("Patients:", nrow(filtered_metadata), "\n")
cat("Group distribution:\n")
print(table(filtered_metadata$PatientGroup))
print(table(filtered_metadata$Classification))
```

---

# 3. Expression Matrix Creation

```{r create-matrix}
# =============================================================================
# PURPOSE: Reshape NPX data from long to wide format for limma
#
# WHY: Olink outputs data in "long" format (one row per protein-sample)
#      limma requires a matrix: rows = proteins, columns = samples
#      This is the standard format for differential expression analysis
# =============================================================================

# Extract the three columns we need
matrix_df <- as.data.frame(cbind(patient_npx$EUSTAR_Number, patient_npx$NPX, patient_npx$OlinkID))
colnames(matrix_df) <- c("EUSTAR_Number", "Value", "Protein")

# Reshape: long → wide (proteins as rows, patients as columns)
wide_data <- tidyr::pivot_wider(matrix_df, names_from = EUSTAR_Number, values_from = Value)
wide_data <- as.data.frame(wide_data)

# Set protein IDs as rownames (limma convention)
rownames(wide_data) <- wide_data$Protein
data <- wide_data[, !(colnames(wide_data) %in% c("NA", "Protein")), drop = FALSE]

# Convert to numeric matrix
# WHY: pivot_wider creates character columns; limma needs numeric
data_matrix <- as.matrix(data)
data_matrix2 <- apply(data_matrix, 2, as.numeric)
rownames(data_matrix2) <- rownames(data_matrix)

# Create gene lookup table (OlinkID → gene symbol)
# WHY: Results are easier to interpret with gene names than Olink IDs
gene_lookup <- patient_npx %>% distinct(OlinkID, Assay)

cat("Expression matrix:", nrow(data_matrix2), "proteins x", ncol(data_matrix2), "patients\n")
```

---

# 4. CRITICAL: Verify Sample Alignment

```{r verify-alignment}
# =============================================================================
# PURPOSE: Ensure metadata rows match expression matrix columns
#
# WHY THIS IS CRITICAL:
# limma assumes row i of the design matrix describes column i of the data matrix.
# If they're misaligned, you're testing the wrong samples against each other!
#
# This is the #1 source of errors in differential expression analysis.
# Always verify alignment before running any statistical model.
# =============================================================================

# Show the mismatch (expected — data come from different sources)
cat("BEFORE alignment:\n")
cat("Matrix columns (first 5):", head(colnames(data_matrix2), 5), "\n")
cat("Metadata EUSTAR (first 5):", head(as.character(filtered_metadata$EUSTAR_Number), 5), "\n")
cat("Aligned:", all(colnames(data_matrix2) == as.character(filtered_metadata$EUSTAR_Number)), "\n\n")

# FIX: Reorder metadata rows to match matrix column order
# match() returns the position in metadata that corresponds to each matrix column
filtered_metadata <- filtered_metadata[match(colnames(data_matrix2),
                                              as.character(filtered_metadata$EUSTAR_Number)), ]

# Verify alignment with stopifnot() — script will halt if alignment fails
stopifnot(all(colnames(data_matrix2) == as.character(filtered_metadata$EUSTAR_Number)))
cat("✓ Alignment verified\n")
```

---

# 5. Analysis A: Binary Group Comparison (Limma)

**Question:** Which proteins differ between Regressors and Non_Regressors?

```{r limma-binary}
# =============================================================================
# PURPOSE: Test for differential protein abundance between groups
#
# WHY LIMMA:
# - Designed for high-dimensional data with small sample sizes
# - Uses empirical Bayes to "borrow strength" across proteins
# - More stable variance estimates than t-tests when n is small
# - Handles covariates (Visit_Year) naturally
#
# MODEL: ~0 + PatientGroup + Visit_Year
# - "~0 + PatientGroup" estimates a mean for each group (no intercept)
# - "+ Visit_Year" adjusts for batch effects from sample collection year
# =============================================================================

# Build design matrix
# WHY ~0 (no intercept): Allows direct group mean estimation
# This makes contrast specification clearer: GroupA - GroupB
design_binary <- model.matrix(~0 + PatientGroup + Visit_Year, data = filtered_metadata)

# Fit linear model to each protein
# For each protein: Expression ~ β1*Regressor + β2*NonRegressor + β3*Year2019 + ...
fit_binary <- lmFit(data_matrix2, design_binary)

# Define contrast: Regressor minus Non_Regressor
# Positive logFC = higher in Regressor, Negative = higher in Non_Regressor
contrast.matrix <- makeContrasts(
  Regressor_vs_NonRegressor = PatientGroupRegressor - PatientGroupNon_Regressor,
  levels = design_binary
)

# Apply contrast and compute moderated statistics
fit_binary2 <- contrasts.fit(fit_binary, contrast.matrix)

# Empirical Bayes moderation
# WHY: Shrinks protein-specific variances toward a common value
# This stabilizes p-values when sample sizes are small
fit_binary2 <- eBayes(fit_binary2)

# Extract results for all proteins
# number = Inf: return all proteins, not just top hits
# adjust.method = "BH": Benjamini-Hochberg FDR correction
binary_results <- topTable(fit_binary2, coef = 1, number = Inf, adjust.method = "BH")

# Add gene symbols for interpretability
binary_results$OlinkID <- rownames(binary_results)
binary_results <- binary_results %>% left_join(gene_lookup, by = "OlinkID")

# Summary statistics
cat("=== BINARY COMPARISON: Regressor vs Non_Regressor ===\n")
cat("FDR < 0.05:", sum(binary_results$adj.P.Val < 0.05), "\n")
cat("FDR < 0.1:", sum(binary_results$adj.P.Val < 0.1), "\n")
cat("FDR < 0.2:", sum(binary_results$adj.P.Val < 0.2), "\n")
cat("Nominal p < 0.05:", sum(binary_results$P.Value < 0.05), "\n")
```

```{r binary-top-hits}
# Display top hits
# logFC: log2 fold change (Regressor vs Non_Regressor)
# AveExpr: average expression across all samples (higher = more abundant)
# P.Value: nominal p-value from moderated t-test
# adj.P.Val: FDR-adjusted p-value (Benjamini-Hochberg)
binary_results %>%
  select(Assay, logFC, AveExpr, P.Value, adj.P.Val) %>%
  head(30)
```

---

# 6. Analysis B: Continuous Correlation with ΔMRSS (Limma)

**Question:** Which proteins predict degree of skin improvement?

```{r limma-continuous}
# =============================================================================
# PURPOSE: Test correlation between protein levels and degree of skin change
#
# WHY CONTINUOUS INSTEAD OF BINARY:
# - Binary (Regressor vs Non_Regressor) treats all improvers equally
# - Continuous analysis uses the actual ΔMRSS value
# - A patient who improved by 15 points contributes more signal than one who improved by 5
# - This can be MORE POWERFUL when the true relationship is dose-dependent
#
# MODEL: ~MRSS_signed + Visit_Year
# - MRSS_signed: continuous predictor (BL - FUP; positive = improvement)
# - Visit_Year: adjustment for batch effects
#
# INTERPRETATION OF logFC:
# - logFC = change in NPX per 1-unit increase in MRSS_signed
# - Positive logFC: higher protein → more improvement
# - Negative logFC: higher protein → more worsening
# =============================================================================

# Build design matrix with continuous predictor
# Note: ~MRSS_signed (with intercept) vs ~0+MRSS_signed (without)
# With intercept is standard for continuous regression
design_continuous <- model.matrix(~MRSS_signed + Visit_Year, data = filtered_metadata)

# Fit linear model
fit_continuous <- lmFit(data_matrix2, design_continuous)
fit_continuous <- eBayes(fit_continuous)

# Extract results for the MRSS_signed coefficient specifically
# coef = "MRSS_signed" extracts just the slope, not the intercept or year effects
continuous_results <- topTable(fit_continuous, coef = "MRSS_signed", number = Inf, adjust.method = "BH")
continuous_results$OlinkID <- rownames(continuous_results)
continuous_results <- continuous_results %>% left_join(gene_lookup, by = "OlinkID")

# Summary
cat("=== CONTINUOUS ANALYSIS: Correlation with ΔMRSS ===\n")
cat("MRSS_signed range:", range(filtered_metadata$MRSS_signed), "\n")
cat("FDR < 0.05:", sum(continuous_results$adj.P.Val < 0.05), "\n")
cat("FDR < 0.1:", sum(continuous_results$adj.P.Val < 0.1), "\n")
cat("FDR < 0.2:", sum(continuous_results$adj.P.Val < 0.2), "\n")
cat("FDR < 0.3:", sum(continuous_results$adj.P.Val < 0.3), "\n")
cat("Nominal p < 0.05:", sum(continuous_results$P.Value < 0.05), "\n")
```

```{r continuous-top-hits}
# Display top hits
# logFC here = NPX change per 1-unit ΔMRSS (not per group like binary analysis)
continuous_results %>%
  select(Assay, logFC, AveExpr, P.Value, adj.P.Val) %>%
  head(30)
```

```{r continuous-effect-size}
# =============================================================================
# PURPOSE: Convert logFC to biologically meaningful effect sizes
#
# WHY: logFC from continuous analysis is "per 1-unit MRSS change"
#      To understand total effect, multiply by the range of MRSS changes observed
#
# EXAMPLE: If logFC = 0.02 and MRSS range is -10 to +20 (span = 30)
#          Total effect = 0.02 × 30 = 0.6 NPX units
#          Fold change = 2^0.6 ≈ 1.5× (50% difference across the range)
# =============================================================================

mrss_span <- diff(range(filtered_metadata$MRSS_signed))

continuous_results %>%
  filter(P.Value < 0.01) %>%
  mutate(
    # Total NPX change across the full range of MRSS values
    total_effect_NPX = logFC * mrss_span,
    # Convert to fold change (NPX is log2 scale)
    fold_change = round(2^total_effect_NPX, 2)
  ) %>%
  select(Assay, logFC, total_effect_NPX, fold_change, P.Value, adj.P.Val) %>%
  head(20)
```

---

# 7. Analysis C: Regressor vs Progressor (Binary)

**Excludes Stable patients** — compares only the extremes.

```{r subset-prog-reg}
# =============================================================================
# PURPOSE: Compare only Regressors vs Progressors (exclude Stable)
#
# WHY EXCLUDE STABLE:
# - Stable patients neither improved nor worsened significantly
# - They may represent a biologically distinct group OR measurement noise
# - Comparing "improvers vs worseners" directly maximizes contrast
# - This is a more stringent test: can we distinguish true opposites?
#
# TRADEOFF: Smaller sample size (lose Stable patients)
#           vs cleaner biological distinction
# =============================================================================

# Subset to Regressor and Progressor only
meta_prog_reg <- filtered_metadata %>%
  filter(Classification %in% c("Regressor", "Progressor")) %>%
  # Set Progressor as reference level (positive logFC = higher in Regressor)
  mutate(Classification = factor(Classification, levels = c("Progressor", "Regressor")))

cat("Regressor vs Progressor subset:\n")
print(table(meta_prog_reg$Classification))
cat("Total n =", nrow(meta_prog_reg), "\n")

# Subset the expression matrix to match
keep_prog_reg <- as.character(meta_prog_reg$EUSTAR_Number)
data_prog_reg <- data_matrix2[, colnames(data_matrix2) %in% keep_prog_reg]

# CRITICAL: Re-align metadata to subsetted matrix
meta_prog_reg <- meta_prog_reg[match(colnames(data_prog_reg),
                                      as.character(meta_prog_reg$EUSTAR_Number)), ]
stopifnot(all(colnames(data_prog_reg) == as.character(meta_prog_reg$EUSTAR_Number)))
cat("✓ Alignment verified\n")
```

```{r limma-binary-prog-reg}
# =============================================================================
# PURPOSE: Limma differential expression — Regressor vs Progressor
#
# Same approach as Analysis A, but on the subset without Stable patients
# =============================================================================

# Design matrix with Visit_Year adjustment
design_prog_reg <- model.matrix(~0 + Classification + Visit_Year, data = meta_prog_reg)

# Fit linear model
fit_prog_reg <- lmFit(data_prog_reg, design_prog_reg)

# Contrast: Regressor vs Progressor
contrast_prog_reg <- makeContrasts(
  Regressor_vs_Progressor = ClassificationRegressor - ClassificationProgressor,
  levels = design_prog_reg
)

fit_prog_reg2 <- contrasts.fit(fit_prog_reg, contrast_prog_reg)
fit_prog_reg2 <- eBayes(fit_prog_reg2)

# Extract results
binary_prog_reg_results <- topTable(fit_prog_reg2, coef = 1, number = Inf, adjust.method = "BH")
binary_prog_reg_results$OlinkID <- rownames(binary_prog_reg_results)
binary_prog_reg_results <- binary_prog_reg_results %>% left_join(gene_lookup, by = "OlinkID")

# Summary
cat("=== BINARY COMPARISON: Regressor vs Progressor ===\n")
cat("FDR < 0.05:", sum(binary_prog_reg_results$adj.P.Val < 0.05), "\n")
cat("FDR < 0.1:", sum(binary_prog_reg_results$adj.P.Val < 0.1), "\n")
cat("FDR < 0.2:", sum(binary_prog_reg_results$adj.P.Val < 0.2), "\n")
cat("Nominal p < 0.05:", sum(binary_prog_reg_results$P.Value < 0.05), "\n")
```

```{r binary-prog-reg-top-hits}
# Top hits
binary_prog_reg_results %>%
  select(Assay, logFC, AveExpr, P.Value, adj.P.Val) %>%
  head(30)
```

---

# 8. Analysis D: Continuous Correlation (Regressor vs Progressor Only)

**Question:** Among patients who definitively improved or worsened, which proteins correlate with degree of change?

```{r limma-continuous-prog-reg}
# =============================================================================
# PURPOSE: Continuous correlation on the Regressor/Progressor subset
#
# WHY THIS ANALYSIS:
# - Combines the benefits of continuous (uses full ΔMRSS range) with
#   the cleaner comparison (excludes ambiguous Stable patients)
# - Stable patients cluster around ΔMRSS = 0, potentially adding noise
# - Removing them may increase power to detect real correlations
#
# EXPECTED DIFFERENCE FROM ANALYSIS B:
# - Wider MRSS_signed range (no values near 0 from Stable patients)
# - Potentially stronger correlations if Stable were adding noise
# - Potentially weaker if sample size reduction hurts more than noise removal helps
# =============================================================================

# Design matrix with continuous predictor
design_cont_prog_reg <- model.matrix(~MRSS_signed + Visit_Year, data = meta_prog_reg)

# Fit and moderate
fit_cont_prog_reg <- lmFit(data_prog_reg, design_cont_prog_reg)
fit_cont_prog_reg <- eBayes(fit_cont_prog_reg)

# Extract results for MRSS_signed coefficient
continuous_prog_reg_results <- topTable(fit_cont_prog_reg, coef = "MRSS_signed", number = Inf, adjust.method = "BH")
continuous_prog_reg_results$OlinkID <- rownames(continuous_prog_reg_results)
continuous_prog_reg_results <- continuous_prog_reg_results %>% left_join(gene_lookup, by = "OlinkID")

# Summary
cat("=== CONTINUOUS ANALYSIS: Regressor vs Progressor Only ===\n")
cat("MRSS_signed range:", range(meta_prog_reg$MRSS_signed), "\n")
cat("FDR < 0.05:", sum(continuous_prog_reg_results$adj.P.Val < 0.05), "\n")
cat("FDR < 0.1:", sum(continuous_prog_reg_results$adj.P.Val < 0.1), "\n")
cat("FDR < 0.2:", sum(continuous_prog_reg_results$adj.P.Val < 0.2), "\n")
cat("Nominal p < 0.05:", sum(continuous_prog_reg_results$P.Value < 0.05), "\n")
```

```{r continuous-prog-reg-top-hits}
# Top hits
continuous_prog_reg_results %>%
  select(Assay, logFC, AveExpr, P.Value, adj.P.Val) %>%
  head(30)
```

```{r}
## Boxplot (Regressor vs Progressor Only)

plot_protein_boxplot_RP <- function(protein_id, gene_name = NULL) {
  # PURPOSE: Boxplot for Regressor vs Progressor comparison (excludes Stable)
  # Shows Wilcoxon p-value for binary comparison

  # Auto-lookup gene name if not provided
  if (is.null(gene_name)) {
    gene_name <- gene_lookup$Assay[gene_lookup$OlinkID == protein_id]
    if (length(gene_name) == 0 || is.na(gene_name)) gene_name <- protein_id
  }

  # Use R vs P subset data
  plot_data <- data.frame(
    Expression = data_prog_reg[protein_id, ],
    Group = meta_prog_reg$Classification
  )

  p <- ggplot(plot_data, aes(x = Group, y = Expression, fill = Group)) +
    geom_violin(alpha = 0.3, width = 0.8, color = NA, trim = TRUE) +
    geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.8) +
    geom_jitter(width = 0.1, size = 2, alpha = 0.6) +
    stat_compare_means(method = "wilcox.test", label = "p.format",
                       label.x.npc = "center", label.y.npc = 0.95) +
    scale_fill_manual(values = c("Progressor" = "#AF3B6E", "Regressor" = "#48639C")) +
    labs(title = gene_name, x = NULL, y = "NPX") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
          legend.position = "none")

  return(p)
}
## Scatterplot (Regressor vs Progressor Only)

plot_protein_scatter_RP <- function(protein_id, gene_name = NULL) {
  # PURPOSE: Scatterplot for R vs P subset showing correlation with ΔMRSS

  # Auto-lookup gene name
  if (is.null(gene_name)) {
    gene_name <- gene_lookup$Assay[gene_lookup$OlinkID == protein_id]
    if (length(gene_name) == 0 || is.na(gene_name)) gene_name <- protein_id
  }

  # Get ADJUSTED p-value from R vs P continuous analysis
  cont_row <- continuous_prog_reg_results %>% filter(OlinkID == protein_id)
  pval_text <- if (nrow(cont_row) > 0) sprintf("FDR = %.2g", cont_row$adj.P.Val) else ""

  plot_data <- data.frame(
    Expression = data_prog_reg[protein_id, ],
    MRSS_signed = meta_prog_reg$MRSS_signed,
    Group = meta_prog_reg$Classification
  )

  p <- ggplot(plot_data, aes(x = MRSS_signed, y = Expression)) +
    geom_point(aes(color = Group), size = 2.5, alpha = 0.7) +
    geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 0.8) +
    scale_color_manual(values = c("Progressor" = "#AF3B6E", "Regressor" = "#48639C")) +
    labs(title = gene_name, subtitle = pval_text,
         x = "ΔMRSS (positive = improvement)", y = "NPX") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
          plot.subtitle = element_text(hjust = 0.5, color = "gray40"))

  return(p)
}


```

---

# 9. Sensitivity Analysis: Time-Balanced Subset

```{r check-confounding}
# =============================================================================
# PURPOSE: Check for confounding between PatientGroup and Visit_Year
#
# WHY THIS MATTERS:
# - If all Progressors were sampled in 2022-2023 and all Regressors in 2014-2021,
#   any "group difference" might actually be a batch effect
# - This table reveals the extent of confounding
# =============================================================================

cat("=== CONFOUNDING CHECK: Group × Visit_Year ===\n")
print(table(filtered_metadata$PatientGroup, filtered_metadata$Visit_Year))
```

```{r sensitivity-analysis}
# =============================================================================
# PURPOSE: Sensitivity analysis restricted to time-balanced years
#
# WHY: The full dataset has Visit_Year confounded with PatientGroup.
#      Some years have only one group represented.
#      By restricting to years with BOTH groups, we can:
#      1. Test if results hold without batch adjustment
#      2. Identify proteins that are robust vs potentially batch-driven
#
# INTERPRETATION:
# - Proteins significant in BOTH full and sensitivity analyses = robust
# - Proteins significant only in full analysis = may be batch effects
# - Proteins significant only in sensitivity = may be real but underpowered in full
#
# LIMITATION: Smaller sample size reduces power
# =============================================================================

# Restrict to years where both groups are represented
overlapping_years <- c("2014", "2019", "2020", "2021")

meta_sensitivity <- filtered_metadata %>%
  filter(Visit_Year %in% overlapping_years)

cat("\nSensitivity analysis sample:\n")
print(table(meta_sensitivity$PatientGroup, meta_sensitivity$Visit_Year))
cat("Total n =", nrow(meta_sensitivity), "\n")

# Subset expression matrix
keep_samples <- as.character(meta_sensitivity$EUSTAR_Number)
data_sensitivity <- data_matrix2[, colnames(data_matrix2) %in% keep_samples]

# Align metadata
meta_sensitivity <- meta_sensitivity[match(colnames(data_sensitivity),
                                            as.character(meta_sensitivity$EUSTAR_Number)), ]
stopifnot(all(colnames(data_sensitivity) == as.character(meta_sensitivity$EUSTAR_Number)))

# Run continuous model WITHOUT Visit_Year adjustment
# WHY: Within overlapping years, groups are more balanced
#      This tests if results hold without needing batch correction
design_sensitivity <- model.matrix(~MRSS_signed, data = meta_sensitivity)
fit_sensitivity <- lmFit(data_sensitivity, design_sensitivity)
fit_sensitivity <- eBayes(fit_sensitivity)

sensitivity_results <- topTable(fit_sensitivity, coef = "MRSS_signed", number = Inf, adjust.method = "BH")
sensitivity_results$OlinkID <- rownames(sensitivity_results)
sensitivity_results <- sensitivity_results %>% left_join(gene_lookup, by = "OlinkID")

cat("\n=== SENSITIVITY ANALYSIS RESULTS ===\n")
cat("FDR < 0.05:", sum(sensitivity_results$adj.P.Val < 0.05), "\n")
cat("FDR < 0.1:", sum(sensitivity_results$adj.P.Val < 0.1), "\n")
cat("Nominal p < 0.05:", sum(sensitivity_results$P.Value < 0.05), "\n")
```

```{r sensitivity-top-hits}
# Top hits from sensitivity analysis
sensitivity_results %>%
  select(Assay, logFC, P.Value, adj.P.Val) %>%
  head(20)
```

---

# 10. Compare Results Across All Analyses

```{r compare-analyses}
# =============================================================================
# PURPOSE: Compare each candidate protein across all 5 analyses
#
# WHY: A protein that is significant in only ONE analysis may be:
# - A batch effect (significant only when not adjusting properly)
# - Driven by a few outliers in one subset
# - A true finding that's underpowered in other analyses
#
# ROBUST FINDINGS should show:
# - Consistent direction (logFC same sign) across analyses
# - Significant (or trending) p-values in multiple analyses
# - Survival in sensitivity analysis = not a batch effect
# =============================================================================

compare_protein <- function(protein_name) {
  binary <- binary_results %>% filter(Assay == protein_name) %>%
    select(Assay, logFC, P.Value, adj.P.Val) %>% mutate(Analysis = "Binary (R vs NR)")

  continuous <- continuous_results %>% filter(Assay == protein_name) %>%
    select(Assay, logFC, P.Value, adj.P.Val) %>% mutate(Analysis = "Continuous (all)")

  binary_pr <- binary_prog_reg_results %>% filter(Assay == protein_name) %>%
    select(Assay, logFC, P.Value, adj.P.Val) %>% mutate(Analysis = "Binary (R vs P)")

  continuous_pr <- continuous_prog_reg_results %>% filter(Assay == protein_name) %>%
    select(Assay, logFC, P.Value, adj.P.Val) %>% mutate(Analysis = "Continuous (R vs P)")

  sensitivity <- sensitivity_results %>% filter(Assay == protein_name) %>%
    select(Assay, logFC, P.Value, adj.P.Val) %>% mutate(Analysis = "Sensitivity")

  bind_rows(binary, continuous, binary_pr, continuous_pr, sensitivity)
}

# Check candidate proteins (add your proteins of interest here)
candidates <- c("MMP1", "SERPINE2", "TIMP1", "TGFB1")

cat("=== CANDIDATE PROTEIN COMPARISON ===\n\n")
for (protein in candidates) {
  cat(protein, ":\n")
  print(compare_protein(protein))
  cat("\n")
}
```

---

# 11. Healthy Controls for Visualization

```{r prepare-hc}
# =============================================================================
# PURPOSE: Prepare healthy control (HC) samples for plotting
#
# WHY INCLUDE HC:
# - Provides biological context: are patient levels abnormal?
# - Helps visualize whether Regressors normalize toward healthy levels
# - NOT included in statistical models (different context, no MRSS)
# =============================================================================

# Extract HC samples (identified by sample ID pattern)
hc_npx <- npx_data %>%
  filter(grepl("^HC-OL", SampleID))

hc_sample_ids <- unique(hc_npx$SampleID)
cat("Found", length(hc_sample_ids), "HC samples\n")

# Create HC expression matrix (same format as patient matrix)
hc_matrix_df <- hc_npx %>%
  select(SampleID, NPX, OlinkID) %>%
  pivot_wider(names_from = SampleID, values_from = NPX, id_cols = OlinkID)

hc_matrix <- as.data.frame(hc_matrix_df)
rownames(hc_matrix) <- hc_matrix$OlinkID
hc_matrix <- hc_matrix[, -1]
hc_matrix <- apply(as.matrix(hc_matrix), 2, as.numeric)
rownames(hc_matrix) <- hc_matrix_df$OlinkID
```

```{r combine-for-plotting}
# =============================================================================
# PURPOSE: Create combined dataset for visualization with batch adjustment
#
# WHY ADJUST FOR VISIT_YEAR:
# - Patients were collected over multiple years (potential batch effects)
# - HC samples collected at one timepoint (no year adjustment possible)
# - For fair comparison, we remove estimated Visit_Year effects from patients
#
# HOW: Subtract the fitted Visit_Year coefficients from patient NPX values
# =============================================================================

# Extract Visit_Year effect from the fitted model
visit_cols <- grep("^Visit_Year", colnames(design_binary))
visit_effect <- fit_binary$coefficients[, visit_cols, drop = FALSE] %*% t(design_binary[, visit_cols, drop = FALSE])

# Subtract year effects from patient data
adjusted_npx <- data_matrix2 - visit_effect

# Combine patient (adjusted) and HC (raw) matrices
common_proteins <- intersect(rownames(adjusted_npx), rownames(hc_matrix))
combined_matrix <- cbind(
  adjusted_npx[common_proteins, ],
  hc_matrix[common_proteins, ]
)

# Create combined metadata
combined_metadata <- bind_rows(
  filtered_metadata %>%
    select(SampleID = EUSTAR_Number, PatientGroup) %>%
    mutate(SampleID = as.character(SampleID)),
  data.frame(SampleID = hc_sample_ids, PatientGroup = "Healthy")
) %>%
  # Set factor order for plotting (Healthy first as reference)
  mutate(PatientGroup = factor(PatientGroup, levels = c("Healthy", "Non_Regressor", "Regressor")))

# Align metadata to matrix
combined_metadata <- combined_metadata[match(colnames(combined_matrix), combined_metadata$SampleID), ]
stopifnot(all(colnames(combined_matrix) == combined_metadata$SampleID))

cat("Combined data:", ncol(combined_matrix), "samples\n")
print(table(combined_metadata$PatientGroup))
```

---

# 12. Plotting Functions

```{r plotting-functions-intro}
# =============================================================================
# PURPOSE: Reusable functions to visualize individual proteins
#
# TWO VISUALIZATION TYPES:
# 1. Boxplot: Shows group distributions (matches binary analysis)
# 2. Scatterplot: Shows correlation with ΔMRSS (matches continuous analysis)
#
# USAGE:
#   plot_protein_boxplot("OID44783")  # By Olink ID
#   plot_protein_boxplot("OID44783", gene_name = "MMP1")  # With custom title
#   plot_protein_scatter("OID44783")
# =============================================================================
```

## Boxplot (Binary Comparison)

```{r plot-function-boxplot}
plot_protein_boxplot <- function(protein_id, gene_name = NULL, show_hc = TRUE) {
  # PURPOSE: Visualize protein expression by group
  # protein_id: Olink ID (e.g., "OID44783")
  # gene_name: Optional label override
  # show_hc: Include healthy controls (uses batch-adjusted data)

  # Auto-lookup gene name if not provided
  if (is.null(gene_name)) {
    gene_name <- gene_lookup$Assay[gene_lookup$OlinkID == protein_id]
    if (length(gene_name) == 0 || is.na(gene_name)) gene_name <- protein_id
  }

  # Select appropriate data (with or without HC)
  if (show_hc) {
    plot_data <- data.frame(
      Expression = combined_matrix[protein_id, ],
      Group = combined_metadata$PatientGroup
    )
    comparisons <- list(c("Regressor", "Non_Regressor"))
  } else {
    plot_data <- data.frame(
      Expression = data_matrix2[protein_id, ],
      Group = filtered_metadata$PatientGroup
    )
    comparisons <- list(c("Regressor", "Non_Regressor"))
  }

  p <- ggplot(plot_data, aes(x = Group, y = Expression, fill = Group)) +
    # Violin shows distribution shape
    geom_violin(alpha = 0.3, width = 0.8, color = NA, trim = TRUE) +
    # Box shows median and IQR
    geom_boxplot(width = 0.2, outlier.shape = NA, alpha = 0.8) +
    # Points show individual samples (critical for small n)
    geom_jitter(width = 0.1, size = 2, alpha = 0.6) +
    # Add Wilcoxon test p-value
    stat_compare_means(method = "wilcox.test", comparisons = comparisons,
                       label = "p.format", tip.length = 0.02) +
    scale_fill_manual(values = c("Healthy" = "#2E8B57",
                                  "Non_Regressor" = "#AF3B6E",
                                  "Regressor" = "#48639C")) +
    labs(title = gene_name, x = NULL, y = "NPX") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          legend.position = "none")

  return(p)
}
```

## Scatterplot (Continuous Correlation)

```{r plot-function-scatter}
plot_protein_scatter <- function(protein_id, gene_name = NULL) {
  # PURPOSE: Visualize protein correlation with ΔMRSS
  # Shows the continuous relationship that limma is modeling

  # Auto-lookup gene name
  if (is.null(gene_name)) {
    gene_name <- gene_lookup$Assay[gene_lookup$OlinkID == protein_id]
    if (length(gene_name) == 0 || is.na(gene_name)) gene_name <- protein_id
  }

  # Get p-value from continuous analysis for annotation
  cont_row <- continuous_results %>% filter(OlinkID == protein_id)
  pval_text <- if (nrow(cont_row) > 0) sprintf("p = %.2g", cont_row$P.Value) else ""

  plot_data <- data.frame(
    Expression = data_matrix2[protein_id, ],
    MRSS_signed = filtered_metadata$MRSS_signed,
    Group = filtered_metadata$PatientGroup
  )

  p <- ggplot(plot_data, aes(x = MRSS_signed, y = Expression)) +
    # Color points by group to show overlap/separation
    geom_point(aes(color = Group), size = 2.5, alpha = 0.7) +
    # Add regression line (this is what limma is fitting)
    geom_smooth(method = "lm", se = TRUE, color = "black", linewidth = 0.8) +
    scale_color_manual(values = c("Non_Regressor" = "#AF3B6E", "Regressor" = "#48639C")) +
    labs(title = gene_name, subtitle = pval_text,
         x = "ΔMRSS (positive = improvement)", y = "NPX") +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5, face = "bold"),
          plot.subtitle = element_text(hjust = 0.5, color = "gray40"))

  return(p)
}
```

---

# 13. Volcano Plot

```{r volcano-plot, fig.width=8, fig.height=6}
# =============================================================================
# PURPOSE: Visualize all proteins simultaneously
#
# VOLCANO PLOT ANATOMY:
# - X-axis: Effect size (log2 fold change)
# - Y-axis: Statistical significance (-log10 p-value)
# - Upper corners: Large effect + significant (candidates)
# - Middle: Small effect (even if significant, may not be meaningful)
#
# COLOR CODING:
# - Red: FDR < 0.1 (multiple testing corrected)
# - Orange: Nominal p < 0.05 (uncorrected)
# - Gray: Not significant
# =============================================================================

binary_results <- binary_results %>%
  mutate(Significance = case_when(
    adj.P.Val < 0.1 ~ "FDR < 0.1",
    P.Value < 0.05 ~ "Nominal",
    TRUE ~ "NS"
  ))

volcano <- ggplot(binary_results, aes(x = logFC, y = -log10(P.Value))) +
  geom_point(aes(color = Significance), alpha = 0.6, size = 2) +
  # Horizontal line: nominal significance threshold
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "gray50") +
  # Vertical lines: effect size threshold (|logFC| > 0.5 = 1.4-fold)
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed", color = "gray50") +
  # Label top hits
  geom_text_repel(
    data = binary_results %>% filter(P.Value < 0.01),
    aes(label = Assay), size = 3, max.overlaps = 20
  ) +
  scale_color_manual(values = c("FDR < 0.1" = "#E41A1C", "Nominal" = "#FF7F00", "NS" = "gray70")) +
  labs(title = "Volcano Plot: Regressor vs Non_Regressor",
       x = "log2 Fold Change", y = "-log10(P-value)") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

# Save volcano as PDF
ggsave(file.path(output_dir, "volcano_Regressor_vs_NonRegressor.pdf"), volcano, width = 8, height = 6)

print(volcano)
```

---

# 14. Filter by Average Expression (Optional)

```{r filter-by-expression}
# =============================================================================
# PURPOSE: Reduce multiple testing burden by filtering low-abundance proteins
#
# RATIONALE:
# - Low-abundance proteins (low AveExpr) are measured with more noise
# - They contribute to multiple testing correction but rarely yield true signals
# - Filtering them BEFORE p-value adjustment can improve FDR for remaining proteins
#
# THRESHOLD CHOICE (AveExpr > 1):
# - NPX is log2 scale, so AveExpr = 1 means ~2-fold above background
# - This is a conservative filter; adjust based on your data distribution
#
# IMPORTANT: Must recalculate adjusted p-values after filtering!
# The original adj.P.Val was computed across ALL proteins.
# =============================================================================

continuous_filtered <- continuous_results %>%
  filter(AveExpr > 0)

cat("Before filter:", nrow(continuous_results), "proteins\n")
cat("After filter (AveExpr > 1):", nrow(continuous_filtered), "proteins\n")

# CRITICAL: Recalculate FDR on the filtered subset
continuous_filtered$adj.P.Val_filtered <- p.adjust(continuous_filtered$P.Value, method = "BH")

cat("\nTop hits after filtering:\n")
continuous_filtered %>%
  select(Assay, logFC, AveExpr, P.Value, adj.P.Val_filtered) %>%
  head(20)
```

---

# 15. Export Results

```{r export}
# =============================================================================
# PURPOSE: Save all results for downstream use
#
# OUTPUT FILES:
# 1. Binary Regressor vs Non_Regressor: includes Stable in Non_Regressor
# 2. Continuous (all): correlation with ΔMRSS, all patients
# 3. Binary Regressor vs Progressor: excludes Stable
# 4. Continuous (R vs P): correlation, excludes Stable
# 5. Sensitivity: time-balanced subset, no batch adjustment
#
# COLUMNS IN OUTPUT:
# - logFC: log2 fold change (interpret based on analysis type)
# - AveExpr: average expression (higher = more abundant/reliable)
# - t: moderated t-statistic
# - P.Value: nominal p-value
# - adj.P.Val: FDR-adjusted p-value (Benjamini-Hochberg)
# - B: log-odds of differential expression
# - OlinkID: Olink protein identifier
# - Assay: Gene symbol
# =============================================================================

write.csv(binary_results, file.path(output_dir, "results_binary_Regressor_vs_NonRegressor.csv"), row.names = FALSE)
write.csv(continuous_results, file.path(output_dir, "results_continuous_MRSS_correlation.csv"), row.names = FALSE)
write.csv(binary_prog_reg_results, file.path(output_dir, "results_binary_Regressor_vs_Progressor.csv"), row.names = FALSE)
write.csv(continuous_prog_reg_results, file.path(output_dir, "results_continuous_Regressor_vs_Progressor.csv"), row.names = FALSE)
write.csv(sensitivity_results, file.path(output_dir, "results_sensitivity_timebalanced.csv"), row.names = FALSE)

cat("Exported:\n")
cat("- results_binary_Regressor_vs_NonRegressor.csv\n")
cat("- results_continuous_MRSS_correlation.csv\n")
cat("- results_binary_Regressor_vs_Progressor.csv\n")
cat("- results_continuous_Regressor_vs_Progressor.csv\n")
cat("- results_sensitivity_timebalanced.csv\n")
```

---

# Session Info

```{r session-info}
sessionInfo()
```
